### 2022-10-28 Assignments/paper.tex
upload paper and assignments

\documentclass[runningheads]{llncs}

\usepackage[T1]{fontenc}
\usepackage{graphicx}

\begin{document}

\title{Algorithm Engineering}
\subtitle{Exam Assignments}

\author{Sheela Orgler}
\institute{Friedrich Schiller Universität, Jena}

\maketitle

% \begin{abstract}
%
% \end{abstract}

\section{Assignment}

\subsection{Describe how parallelism differs from concurrency}
Parallelism is a subset of concurrency. If a system is parallel it is also concurrent, but not the other way around. A system is considered parallel if two or more tasks are executed simultaneously.
\\
Concurrency only means support for two or more actions at the same time. Parallelism refers to execution of more than one action at the same time.

\subsection{What is fork-join parallelism?}
Fork-join parallelism refers to a concept where we have a master thread that divides into a team of threads. With this concept we only have parallel regions where threads are executed simultaneously. In between these parallel regions the execution is sequential following the master thread. 
\\
We can imagine execution as a line that is followed from starting to end point. With fork-join parallelism this line is our master thread. Following our master thread at some point the thread can divide into two or more threads. These threads are joined after some time to our master thread. In these sections where we have two or more threads execution will be parallel and after these sections sequential.

\subsection {Chapter 1 – Computer Systems: A Programmer’s Perspective}

\paragraph{Discuss one thing - Caches}
As discussed in Chapter 1 of the book caches are helpful to deal with the processor-memory gap.

The processor can read data from the register file, which is within the CPU almost 100 times faster than from memory. The gap between processor and memory is continuously growing.

Cache memory is used to bridge the gap. It is used to temporarily store information that will likely be used in the future. There exist several levels of caches -  L1, L2, L3 - which go from smallest to largest.

A cache reduces the acess time to data in memory. Frequently used data and instructions are kept in the cache. Access to data kept there is faster.

Modern desktops, servers and industrial CPUs have at least three independent caches: instruction cache, data cache and the TLB. The instruction cache is used to speed up executable instruction fetches. The data cache speeds up data fetch and store. It is organized into more cache levels:
\begin{itemize}
	\item L1 cache - Primary cache is fast but small and usually embedded in the processor chip (CPU)
	\item L2 cache - Secondary cache is larger and can be embedded in the CPU or on a seperate chip or coprocessor containing a high-speed alternative system bus connectin cache and CPU.
	\item L3 cache - specialized memory to improve L1 and L2. L3 is usually double the speed of DRAM
\end{itemize}
The small caches L1 are backed up by larger and slower caches L2, L3 to address the tradeoff between cache latency and hit rate. The faster cache is generally checked firt. If that cache misses, the next chache is checked and so on, before accessing main memory.

For example the ARM-base Apple M1 CPU has 8 cores: 4 high-performance and four high-efficiency cores. The four high-performance cores have a 192 KiB L1 cache for each of the cores and the four high-efficiency cores only have 128 KiB.

The Translation lookaside buffer is used to speed up the translation from virtual to physical addresses. It is part of the memory management unit and not directly related to CPU caches.

% https://en.wikipedia.org/wiki/CPU_cache#First_instruction_cache
% https://www.techtarget.com/searchstorage/definition/cache-memory

\subsection {Paper – There’s plenty of room at the Top: What will drive computer performance after Moore’s law?}

\paragraph{Explain figure “Performance gains after Moore’s law ends”}
The figure illustrates the “Top” and “Bottom” referring to computer performance gains. The “Bottom” refers to the miniaturization of computer components seen in the last decades. \\
Due to physical limits the opportunities for gains at the bottom will slowly come to an end. Nevertheless, there are still opportunities for growth at the “Top”. The “Top” shows the three aspects where growth can be expected: software, algorithms, hardware architecture. These aspects are divided into technology, opportunity and examples all looking into the growth opportunities in the specific area.
\\
Growth opportunities at the “Top”:
\begin{itemize}
	\item making software more efficient by performance engineering
	\item minimizing the time it takes to run and not the development time
	\item an increasing number of processor cores running parallel
	\item reengineer modularity to obtain performance gains
\end{itemize}

\section{Assignment}

\subsection{What causes false sharing?}

\subsection{How do mutual exclusion constructs prevent race conditions?}

\subsection{Explain the differences betweent static and dynamic schedules in OpenMP.}

\subsection{What can we do if we've found a solution while running a parallel for loop in OpenMP, but still have many iterations left?}

\subsection{Coding warmup slide 22: Explain in your own words how std::atomic::compare\_exchange\_weak work}

%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
% \bibliographystyle{splncs04}
% \bibliography{mybibliography}
%
\begin{thebibliography}{8}

%\bibitem{ref_book1}
%Arpaci-Dusseau, Remzi H and Arpaci-Dusseau, Andrea C.: Operating systems: Three easy pieces. Arpaci-Dusseau Books LLCr,
 %(2018)

\bibitem{ref_url1}
Armv8-A Reference Manual pdf, pages 30-70, pages 1708-1808 \\ \url{https://developer.arm.com/documentation/ddi0487/ga}.

\bibitem{ref_url2}
Armv8-A Address Translation pdf, \url{https://developer.arm.com/documentation/100940/0101/}.

\end{thebibliography}
\end{document}

### 2022-10-28 01_Code/hello.cpp

@@ -1,0 +1,14 @@
+#include <iostream>
+#include "omp.h" // required for library routines
+
+using namespace std;
+
+int main()
+{
+#pragma omp parallel // compiler directive
+  {
+    // omp_get_thread_num() is an OpenMP library routine
+    auto thread_id = omp_get_thread_num();
+    cout << "Hello from thread " << thread_id << endl;
+  }
+}

### 2022-10-28 01_Code/pi_monte_carlo.cpp

@@ -1,0 +1,34 @@
+#include <iostream>
+#include <omp.h>
+#include <random>
+
+using namespace std;
+
+int main()
+{
+  int seed = 0;
+  default_random_engine re{seed};
+  uniform_real_distribution<double> zero_to_one{0.0, 1.0};
+
+  int n = 100000000;                 // amount of points to generate
+  int counter = 0;                   // counter for points in the first quarter of a unit circle
+  auto start_time = omp_get_wtime(); // omp_get_wtime() is an OpenMP library routine
+
+  // compute n points and test if they lie within the first quadrant of a unit circle
+  for (int i = 0; i < n; ++i)
+  {
+    auto x = zero_to_one(re); // generate random number between 0.0 and 1.0
+    auto y = zero_to_one(re); // generate random number between 0.0 and 1.0
+    if (x * x + y * y <= 1.0)
+    { // if the point lies in the first quadrant of a unit circle
+      ++counter;
+    }
+  }
+
+  auto run_time = omp_get_wtime() - start_time;
+  auto pi = 4 * (double(counter) / n);
+
+  cout << "pi: " << pi << endl;
+  cout << "run_time: " << run_time << " s" << endl;
+  cout << "n: " << n << endl;
+}

### 2022-10-28 01_Code/pi_numerical_integration.cpp

@@ -1,0 +1,22 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int num_steps = 100000000; // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double sum = 0.0; // for summing up all heights of rectangles
+
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+  for (int i = 0; i < num_steps; i++) {
+    double x = (i + 0.5) * width; // midpoint
+    sum = sum + (1.0 / (1.0 + x * x)); // add new height of a rectangle
+  }
+  double pi = sum * 4 * width; // compute pi
+  double run_time = omp_get_wtime() - start_time;
+
+  cout << "pi with " << num_steps << " steps is " << setprecision(17)
+       << pi << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-10-28 01_Code/pi_openmp_monte_carlo.cpp

@@ -1,0 +1,45 @@
+#include <iostream>
+#include <omp.h>
+#include <random>
+
+using namespace std;
+
+int main()
+{
+    int seed = 0;
+    default_random_engine re{seed};
+    uniform_real_distribution<double> zero_to_one{0.0, 1.0};
+
+    int n = 100000000; // amount of points to generate
+    int counter = 0;   // counter for points in the first quarter of a unit circle
+
+    auto start_time = omp_get_wtime(); // omp_get_wtime() is an OpenMP library routine
+    omp_set_num_threads(4);
+
+#pragma omp parallel
+    {
+        int num_threads = omp_get_num_threads();
+        int thread_id = omp_get_thread_num();
+        int counter_local = 0;
+
+        // compute n points and test if they lie within the first quadrant of a unit circle
+        for (int i = thread_id; i < n; i += num_threads)
+        {
+            auto x = zero_to_one(re); // generate random number between 0.0 and 1.0
+            auto y = zero_to_one(re); // generate random number between 0.0 and 1.0
+            if (x * x + y * y <= 1.0)
+            { // if the point lies in the first quadrant of a unit circle
+                ++counter_local;
+            }
+        }
+#pragma omp atomic
+        counter += counter_local;
+    }
+
+    auto run_time = omp_get_wtime() - start_time;
+    auto pi = 4 * (double(counter) / n);
+
+    cout << "pi: " << pi << endl;
+    cout << "run_time: " << run_time << " s" << endl;
+    cout << "n: " << n << endl;
+}

### 2022-10-28 02_Code/continue_hack.cpp

@@ -1,0 +1,33 @@
+#include <omp.h>
+#include <atomic>
+#include <iostream>
+
+using namespace std;
+
+bool is_solution(int number) {  // test if number solves the problem
+  for (volatile int i = 10000000; i--;) {}  // mock computation
+  return number > 133 && number < 140;
+}
+
+int main() {
+  constexpr int biggest_possible_number = 10000;
+  // To avoid undefined code, it is necessary to use std::atomic for variables 
+  // where race conditions may happen.
+  // https://databasearchitects.blogspot.com/2020/10/c-concurrency-model-on-x86-for-dummies.html
+  atomic<bool> solution_found(false);  // if true than we found the solution
+  atomic<int> final_solution(INT32_MAX);
+  const double start = omp_get_wtime();
+
+#pragma omp parallel for schedule(dynamic)  // start parallel region
+  for (int i = 0; i < biggest_possible_number; ++i) {
+    if (solution_found)  // we found the solution, just continue iterating
+      continue;
+    if (is_solution(i)) {     // find some solution, not necessary the smallest
+      solution_found = true;
+      final_solution = i;
+    }
+  }  // end parallel region
+  // check if we've found a solution at all is omitted, you can add the check
+  cout << "The solution is: " << final_solution << endl;
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-10-28 02_Code/continue_hack_min_solution_atomic.cpp

@@ -1,0 +1,35 @@
+#include <iostream>
+#include <atomic>
+#include <omp.h>
+
+using namespace std;
+
+bool is_solution(int number) { // test if number solves the problem
+  for (volatile int i = 10000000; i--;) {} // mock computation
+  return number > 133 && number < 140;
+}
+
+int main() {
+  constexpr int biggest_possible_number = 10000;
+  atomic<int> final_solution(INT32_MAX); // now we use an atomic
+  const double start = omp_get_wtime();
+
+#pragma omp parallel for schedule(dynamic) // start parallel region
+  for (int i = 0; i < biggest_possible_number; ++i) {
+    if (final_solution < i) // we found a smaller solution, just continue iterating
+      continue;
+    if (is_solution(i)) { // one solution found, update final_solution if it's smaller
+      int previous = final_solution.load();
+      while (previous > i && !final_solution.compare_exchange_weak(previous, i)) {}
+      // http://www.cplusplus.com/reference/atomic/atomic/compare_exchange_weak/
+      // atomic.compare_exchange_weak(T& expected, T desired);
+      // 1. Compare a given value expected with the value stored in atomic.
+      // 2. If both values stored in expected and atomic coincide then set atomic to a
+      // given value desired, otherwise write the actual value stored in atomic to expected.
+      // 3. Return true if the swap in 2. was successful, otherwise return false.
+    }
+  } // end parallel region
+  // check if we've found a solution at all is omitted, you can add the check
+  cout << "The solution is: " << final_solution << endl;
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-10-28 02_Code/continue_hack_min_solution_lock.cpp

@@ -1,0 +1,32 @@
+#include <atomic>
+#include <iostream>
+#include <mutex>
+#include <omp.h>
+
+using namespace std;
+
+bool is_solution(int number) { // test if number solves the problem
+  for (volatile int i = 10000000; i--;) {} // mock computation
+  return number > 133 && number < 140;
+}
+
+int main() {
+  constexpr int biggest_possible_number = 10000;
+  atomic<int> final_solution(INT32_MAX);
+  const double start = omp_get_wtime();
+  mutex m; // mutual exclusion construct for updating final_solution
+
+#pragma omp parallel for schedule(dynamic) // we could also use schedule(static, 1)
+  for (int i = 0; i < biggest_possible_number; ++i) {
+    if (final_solution < i) // we found a smaller solution, just continue iterating
+      continue;
+    if (is_solution(i)) {    // one solution found, update final_solution if it's smaller
+      lock_guard<mutex> lg(m); // constructs a lock_guard, locking the given mutex
+      final_solution = min(i, final_solution.load()); // only one thread computes min
+    }
+  }
+  // end parallel region
+  // check if we've found a solution at all is omitted, you can add the check
+  cout << "The smallest solution is: " << final_solution << endl;
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-10-28 02_Code/false_sharing.cpp

@@ -1,0 +1,21 @@
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  const int threads = 4; // number of threads
+  volatile long arr[threads]; // no compiler optimizations on arr
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+#pragma omp parallel num_threads(threads)
+  {
+    const int thread_id = omp_get_thread_num();
+    for (long i = 0; i < 1000000000; ++i) {
+      // each thread uses (probably) the same cache line
+      ++arr[thread_id];
+    }
+  }
+  // cout the byte where arr starts in the cache line
+  cout <<  ((long) &arr[0]) % 64 << endl;
+  cout << omp_get_wtime() - start_time << " seconds\n";
+}

### 2022-10-28 02_Code/no_false_sharing.cpp

@@ -1,0 +1,20 @@
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  const int threads = 4; // number of threads
+  const int PAD = 8; // padding value to avoid false sharing
+  volatile long arr[threads * PAD]; // no compiler optimizations on arr
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+#pragma omp parallel num_threads(threads)
+  {
+    const int thread_id = omp_get_thread_num();
+    for (long i = 0; i < 1000000000; ++i) {
+      // each thread increments a value on a different cache line
+      ++arr[thread_id * PAD];
+    }
+  }
+  cout << omp_get_wtime() - start_time << " seconds\n";
+}

### 2022-10-28 02_Code/pi_openmp_v1.cpp

@@ -1,0 +1,29 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int num_steps = 100000000; // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double sum = 0.0; // for summing up all heights of rectangles
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+  omp_set_num_threads(4); // setting the desired number of threads in the parallel region
+#pragma omp parallel // parallel region starts here
+  { // idea for parallelisation: split up loop iterations between threads
+    int num_threads = omp_get_num_threads(); // used as increment in for loop
+    int thread_id = omp_get_thread_num();
+    double sum_local = 0.0; // for summing up heights locally
+    for (int i = thread_id; i < num_steps; i += num_threads) {
+      double x = (i + 0.5) * width; // midpoint
+      sum_local = sum_local + (1.0 / (1.0 + x * x)); // add new height
+    }
+#pragma omp atomic // only one thread performs the update of the sum
+    sum += sum_local;
+  } // parallel region ends here
+  double pi = sum * 4 * width; // compute pi
+  double run_time = omp_get_wtime() - start_time;
+  cout << "pi with " << num_steps << " steps is " << setprecision(17) << pi
+       << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-10-28 02_Code/pi_slow.cpp

@@ -1,0 +1,25 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int num_steps = 100000000; // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+  double pi = 0.0;
+#pragma omp parallel num_threads(4)
+  {
+    int num_threads = omp_get_num_threads();
+    int thread_id = omp_get_thread_num();
+    for (int i = thread_id; i < num_steps; i += num_threads) {
+      double x = (i + 0.5) * width; // midpoint
+#pragma omp atomic
+      pi = pi + (4.0 / (1.0 + x * x)) * width;
+    }
+  }
+  double run_time = omp_get_wtime() - start_time;
+  cout << "pi with " << num_steps << " steps is " << setprecision(17)
+       << pi << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-10-28 02_Code/pragma_omp_for.cpp

@@ -1,0 +1,27 @@
+#include <iostream>
+#include <omp.h>
+#include <vector>
+using namespace std;
+
+struct image;
+
+image *denoise_image(image *in) { // mock denoise
+  image *out = nullptr;
+  for (volatile int i = 0; i < 20000000; ++i) {}
+  return out;
+}
+
+int main() {
+  int amount_images = 64;
+  vector<image *> images(amount_images, nullptr); // mock images
+  vector<image *> denoised_images(amount_images); // mock output
+  double start = omp_get_wtime();
+#pragma omp parallel
+  {
+#pragma omp for // denoise a bunch of images
+    for (int i = 0; i < amount_images; ++i) {
+      denoised_images[i] = denoise_image(images[i]);
+    }
+  }
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-10-28 02_Code/race_condition.cpp

@@ -1,0 +1,26 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int num_steps = 100000000; // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+  double pi = 0.0;
+#pragma omp parallel num_threads(4)
+  {
+    int num_threads = omp_get_num_threads();
+    int thread_id = omp_get_thread_num();
+    double sum_local = 0.0; // for summing up heights locally
+    for (int i = thread_id; i < num_steps; i += num_threads) {
+      double x = (i + 0.5) * width; // midpoint
+      sum_local = sum_local + (1.0 / (1.0 + x * x)); // add new height
+    }
+    pi += sum_local * 4 * width;
+  }
+  double run_time = omp_get_wtime() - start_time;
+  cout << "pi with " << num_steps << " steps is " << setprecision(17) 
+	    << pi << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-10-28 02_Code/schedule_clause.cpp

@@ -1,0 +1,24 @@
+#include <iostream>
+#include <omp.h>
+#include <sstream>
+
+using namespace std;
+
+int main() {
+#pragma omp parallel num_threads(4)
+  {
+    // each thread has its own stringstream instance
+    stringstream thread_iterations;
+    // task for you: experiment with different schedules
+#pragma omp for schedule(static, 8)
+    for (int i = 0; i < 32; ++i) {
+      // write the iterations of a thread to the stream
+      thread_iterations << i << " ";
+    }
+    // only one thread is allowed to output to the console
+#pragma omp critical
+      cout << "thread " << omp_get_thread_num()
+           << " computed the iterations: "
+           << thread_iterations.str() << endl;
+  }
+}

### 2022-11-02 Assignments/paper.tex
add assignment 2

@@
 \subsection{What causes false sharing?}
+False sharing occurs, when threads on different processors modify variables on the same cache line.
+A cache line is the smallest unit of memory.
+Its length depends on the underlying architecture and is typically 64 and in more recent architectures 128 bytes long.
+\\
+False sharing happens when for example two threads on different processors modify variables on the same cache line.
+The first thread modifies one variable at the beginning of the cache line.
+The second thread tries to modify a variable at the end of the cache line.
+Due to the alteration of the first thread, the cache line is invalidated for the second thread and has to be reloaded.
+\\
+False sharing only happens when variables are changed and not when they are only read. It may lead to a significant performance decrease.
 
 \subsection{How do mutual exclusion constructs prevent race conditions?}
+Race conditions occur when two threads modify the same data.
+Mutual exclusions give one thread exclusive access to the data.
+After execution exclusive access is returned and the next thread can execute.
 
 \subsection{Explain the differences betweent static and dynamic schedules in OpenMP.}
+Static and dynamic schedules differ in how the work \( iterations of the for loop \)) is spread across the threads.
+Static means that it is decided at the beginning and dynamic means that it is decided at runtime.
+Each thread will work on a chunk of values and then take the next chunk that hasn't been worked on by any thread.
+Static schedules perform better for balanced workloads and dynamic schedules for unbalanced workloads, in case the workload varies between different iterations of the for loop.
+The chunk size can be specifies; for static schedules it is one per default.
 
 \subsection{What can we do if we've found a solution while running a parallel for loop in OpenMP, but still have many iterations left?}
+If we've found a solution within a parallel for loop, we can use continue. We don't continue calculating, but quickly iterate through the loop to increase performance.
 
-\subsection{Coding warmup slide 22: Explain in your own words how std::atomic::compare\_exchange\_weak work}
+\subsection{Explain in your own words how std::atomic::compare\_exchange\_weak work}
+strd::atomic::compare\_exchange\_weak  succeeds only if the value of the variable to be updated equals the first provided argument.
+For example if for final\_solution.compare\_exchange\_weak(previous, i) final\_solution equals to previous. Previous is updated with i, if it fails.
 

### 2022-11-02 02_Code/pi_openmp_v2.cpp

@@ -1,0 +1,33 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main()
+{
+    int num_steps = 100000000;                   // amount of rectangles
+    double width = 1.0 / double(num_steps);      // width of a rectangle
+    double sum = 0.0;                            // for summing up all heights of rectangles
+    double start_time = omp_get_wtime();         // wall clock time in seconds
+    omp_set_num_threads(4);                      // setting the desired number of threads in the parallel region
+#pragma omp parallel                             // parallel region starts here
+    {                                            // idea for parallelisation: split up loop iterations between threads
+        int num_threads = omp_get_num_threads(); // used as increment in for loop
+        int thread_id = omp_get_thread_num();
+        double sum_local = 0.0; // for summing up heights locally
+
+#pragma omp for
+        for (int i = 0; i < num_steps; ++i)
+        {
+            double x = (i + 0.5) * width;                  // midpoint
+            sum_local = sum_local + (1.0 / (1.0 + x * x)); // add new height
+        }
+#pragma omp atomic // only one thread performs the update of the sum
+        sum += sum_local;
+    }                            // parallel region ends here
+    double pi = sum * 4 * width; // compute pi
+    double run_time = omp_get_wtime() - start_time;
+    cout << "pi with " << num_steps << " steps is " << setprecision(17) << pi
+         << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-11-02 02_Code/pi_slow.cpp

@@ -4,18 +4,20 @@
 
 using namespace std;
 
-int main() {
-  int num_steps = 100000000; // amount of rectangles
+int main()
+{
+  int num_steps = 100000000;              // amount of rectangles
   double width = 1.0 / double(num_steps); // width of a rectangle
-  double start_time = omp_get_wtime(); // wall clock time in seconds
+  double start_time = omp_get_wtime();    // wall clock time in seconds
   double pi = 0.0;
 #pragma omp parallel num_threads(4)
   {
     int num_threads = omp_get_num_threads();
     int thread_id = omp_get_thread_num();
-    for (int i = thread_id; i < num_steps; i += num_threads) {
+    for (int i = thread_id; i < num_steps; i += num_threads)
+    {
       double x = (i + 0.5) * width; // midpoint
-#pragma omp atomic
+#pragma omp critical
       pi = pi + (4.0 / (1.0 + x * x)) * width;
     }
   }

### 2022-11-02 02_Code/race_condition.cpp

@@ -1,26 +1,32 @@
 #include <iomanip>
 #include <iostream>
 #include <omp.h>
+#include <mutex>
 
 using namespace std;
 
-int main() {
-  int num_steps = 100000000; // amount of rectangles
+int main()
+{
+  int num_steps = 100000000;              // amount of rectangles
   double width = 1.0 / double(num_steps); // width of a rectangle
-  double start_time = omp_get_wtime(); // wall clock time in seconds
+  double start_time = omp_get_wtime();    // wall clock time in seconds
   double pi = 0.0;
+  mutex m;
 #pragma omp parallel num_threads(4)
   {
     int num_threads = omp_get_num_threads();
     int thread_id = omp_get_thread_num();
     double sum_local = 0.0; // for summing up heights locally
-    for (int i = thread_id; i < num_steps; i += num_threads) {
-      double x = (i + 0.5) * width; // midpoint
+    for (int i = thread_id; i < num_steps; i += num_threads)
+    {
+      double x = (i + 0.5) * width;                  // midpoint
       sum_local = sum_local + (1.0 / (1.0 + x * x)); // add new height
     }
+    m.lock();
     pi += sum_local * 4 * width;
+    m.unlock();
   }
   double run_time = omp_get_wtime() - start_time;
-  cout << "pi with " << num_steps << " steps is " << setprecision(17) 
-	    << pi << " in " << setprecision(6) << run_time << " seconds\n";
+  cout << "pi with " << num_steps << " steps is " << setprecision(17)
+       << pi << " in " << setprecision(6) << run_time << " seconds\n";
 }

### 2022-11-02 02_Code/schedule_clause.cpp

@@ -4,21 +4,23 @@
 
 using namespace std;
 
-int main() {
+int main()
+{
 #pragma omp parallel num_threads(4)
   {
     // each thread has its own stringstream instance
     stringstream thread_iterations;
     // task for you: experiment with different schedules
-#pragma omp for schedule(static, 8)
-    for (int i = 0; i < 32; ++i) {
+#pragma omp for schedule(static, 22)
+    for (int i = 0; i < 32; ++i)
+    {
       // write the iterations of a thread to the stream
       thread_iterations << i << " ";
     }
     // only one thread is allowed to output to the console
 #pragma omp critical
-      cout << "thread " << omp_get_thread_num()
-           << " computed the iterations: "
-           << thread_iterations.str() << endl;
+    cout << "thread " << omp_get_thread_num()
+         << " computed the iterations: "
+         << thread_iterations.str() << endl;
   }
 }

### 2022-11-02 02_Code/unbalanced_workload.cpp

@@ -6,24 +6,29 @@ using namespace std;
 
 struct image;
 
-image *denoise_image(image *in, int scale) { // mock denoise
+image *denoise_image(image *in, int scale)
+{ // mock denoise
   image *out = nullptr;
   const int loop_until = 20000 * scale; // unbalanced workload
-  for (volatile int i = 0; i < loop_until; ++i) {}
+  for (volatile int i = 0; i < loop_until; ++i)
+  {
+  }
   return out;
 }
 
-int main() {
+int main()
+{
   int amount_images = 64;
   vector<image *> images(amount_images, nullptr); // mock images
   vector<image *> denoised_images(amount_images); // mock output
   double start = omp_get_wtime();
-  
+
   // #pragma omp parallel for is an "OpenMP shortcut"
-#pragma omp parallel for num_threads(4) // denoise a bunch of images
-  for (int i = amount_images - 1; i > -1; --i) {
+#pragma omp parallel for num_threads(4) schedule(dynamic) // denoise a bunch of images
+  for (int i = amount_images - 1; i > -1; --i)
+  {
     denoised_images[i] = denoise_image(images[i], i * i);
   }
-  
+
   cout << omp_get_wtime() - start << " seconds" << endl;
 }

### 2022-11-02 03_Code/barrier.cpp
add Code_03

@@ -1,0 +1,18 @@
+#include <iostream>
+#include <omp.h>
+#include <sstream>
+
+using namespace std;
+
+int main() {
+#pragma omp parallel
+  {
+    stringstream info;
+    info << "Hello from thread " << omp_get_thread_num() << endl;
+    cout << info.str();
+#pragma omp barrier // move on as soon as all threads printed Hello ...
+    info.str(""); // "clear" stringstream variable
+    info << "Goodbye from thread " << omp_get_thread_num() << endl;
+    cout << info.str();
+  }
+}

### 2022-11-02 03_Code/collapse.cpp

@@ -1,0 +1,17 @@
+int main() {
+  const int N = 10;
+  const int M = 12;
+
+  /**
+   * will form a single loop of length N * M = 120
+   * and then parallelize that
+   *
+   * can be useful for balancing the work
+   */
+#pragma omp parallel for collapse(2)
+  for (int i = 0; i < N; i++) {
+    for (int j = 0; j < M; j++) {
+      // do useful work with i and j
+    }
+  }
+}

### 2022-11-02 03_Code/grid_search.cpp

@@ -1,0 +1,30 @@
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+using namespace std;
+
+// mock k-nearest neighbors classification
+void knn(int k, const string &weight, const string &metric) {
+  // larger k needs more runtime
+  for (volatile int i = 0; i < 10000000 * k; ++i) {}
+  // computing accuracy of classification is omitted here
+#pragma omp critical // output which thread did what
+  cout << "k: " << k << ", weight: " << weight << ", metric: " << metric
+       << ", computed with thread: " << omp_get_thread_num() << endl;
+}
+
+int main() {
+  vector<int> ks{1, 3, 5, 7, 9, 11};
+  vector<string> weights{"uniform", "distance"};
+  vector<string> metrics{"euclidean", "manhattan"};
+  const double start = omp_get_wtime();
+  // parallel grid search for tuning the hyperparameters in knn
+#pragma omp parallel for collapse(3) schedule(dynamic)
+  for (uint64_t i = 0; i < ks.size(); ++i)
+    for (uint64_t j = 0; j < weights.size(); ++j)
+      for (uint64_t k = 0; k < metrics.size(); ++k)
+        knn(ks[i], weights[j], metrics[k]);
+
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-11-02 03_Code/mandelbrot_tuned.cpp

@@ -1,0 +1,56 @@
+#include <complex>
+#include <fstream>
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+using namespace std;
+
+constexpr int width = 4000;
+constexpr int height = width;
+
+int compute_pixel(int x, int y) { // compute the gray value of a pixel
+  // possible further optimization: complex<float> instead of complex<double>, 
+  // accuracy of float is enough here (may increase speed)
+  complex<double> point(2.0 * x / width - 1.5, 2.0 * y / height - 1.0);
+  complex<double> z(0, 0);
+  constexpr int iterations = 100;
+  int nb_iter = 0;
+  // test if point in mandelbrot set
+  while (abs(z) < 2 && nb_iter < iterations) {
+    z = z * z + point;
+    nb_iter++;
+  }
+  // white color (255) if point in mandelbrot set, else a shade of gray
+  return (255 * nb_iter) / iterations;
+}
+
+int main() { // generate mandelbrot pgm (portable graymap)
+  const string image_name = "mandelbrot.pgm";
+  remove(image_name.c_str()); // remove file from disk
+  const double start = omp_get_wtime();
+  vector<string> look_up{256}; // look-up table for strings
+  for (int i = 0; i < 256; ++i) { look_up[i] = to_string(i) + "\n"; } // fill look-up table
+  ofstream image(image_name); // file output stream
+  if (image.is_open()) {
+    image << "P2\n" << width << " " << height << " 255\n"; // write pgm header
+#pragma omp parallel // start parallel region
+    {
+      string buffer; // each thread has it's own string buffer
+      buffer.reserve(width * 4); // reserve enough space to avoid reallocation
+#pragma omp for schedule(dynamic) ordered // we could also use schedule(static, 1)
+      for (int i = 0; i < height; i++) {
+        buffer.clear(); // clear content of buffer
+        for (int j = 0; j < width; j++) {
+          buffer += look_up[compute_pixel(j, i)]; // fill buffer
+        }
+#pragma omp ordered // write rows of image in sequential order
+        image << buffer; // write string buffer to file
+      }
+    } // end parallel region
+    image.close(); // close file output stream
+  } else {
+    cout << "Could not open the file!";
+  }
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-11-02 03_Code/pi_cpp_threads.cpp

@@ -1,0 +1,51 @@
+#include <chrono>
+#include <iomanip>
+#include <iostream>
+#include <mutex>
+#include <thread>
+#include <vector>
+
+using namespace std;
+using Clock = chrono::steady_clock;
+
+// thunk is called from different threads
+void thunk(int num_threads, int thread_id, int num_steps, double width,
+           double &sum, mutex &m) {
+  double sum_local = 0.0; // for summing up heights locally
+  for (int i = thread_id; i < num_steps; i += num_threads) {
+    double x = (i + 0.5) * width; // midpoint
+    sum_local = sum_local + (1.0 / (1.0 + x * x)); // add new height
+  }
+  lock_guard<mutex> lg(m); // avoids race condition on sum
+  sum += sum_local; // update global sum
+}
+
+int main() {
+  int num_steps = 100000000;              // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double sum = 0.0; // for summing up all heights of rectangles
+  auto start = Clock::now();
+  int num_threads = (int) thread::hardware_concurrency(); // amount of logical cores
+  mutex m;
+  vector<thread> threads;
+  threads.reserve(num_threads);
+
+  // start threads
+  for (int i = 1; i < num_threads; ++i) {
+    thread th(thunk, num_threads, i, num_steps, width, ref(sum), ref(m));
+    threads.push_back(move(th));
+  }
+  // run function thunk on master thread
+  thunk(num_threads, 0, num_steps, width, sum, m);
+
+  // join threads
+  for (std::thread &th : threads) {
+    th.join();
+  }
+  
+  double pi = sum * 4 * width; // compute pi
+  double run_time = (Clock::now() - start).count() / 1e9;
+  cout << "pi with " << num_steps << " steps is " << setprecision(17) << pi
+       << " in " << setprecision(4) << run_time << " seconds\n";
+  return 0;
+}

### 2022-11-02 03_Code/pi_openmp.cpp

@@ -1,0 +1,23 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int num_steps = 100000000; // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double sum = 0.0; // for summing up all heights of rectangles
+
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+#pragma omp parallel for reduction(+:sum)
+  for (int i = 0; i < num_steps; i++) {
+    double x = (i + 0.5) * width; // midpoint
+    sum = sum + (1.0 / (1.0 + x * x)); // add new height of a rectangle
+  }
+  double pi = sum * 4 * width; // compute pi
+  double run_time = omp_get_wtime() - start_time;
+
+  cout << "pi with " << num_steps << " steps is " << setprecision(17)
+       << pi << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-11-02 03_Code/private_vs_firstprivate.cpp

@@ -1,0 +1,27 @@
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int x = 10;
+  omp_set_num_threads(4);
+
+  cout << "private(x): ";
+#pragma omp parallel private(x)
+  { // x is an uninitialized local copy in each thread
+    x += omp_get_thread_num(); // (probably) 0 + thread_num
+#pragma omp critical
+    cout << x << " "; // (probably) something like: 3 2 1 0
+  }
+
+  cout << "\nfirstprivate(x): ";
+#pragma omp parallel firstprivate(x)
+  { // x is an initialized local one-to-one copy in each thread
+    x += omp_get_thread_num(); // 10 + thread_num
+#pragma omp critical
+    cout << x << " "; // something like: 10 12 11 13
+  }
+
+  cout << "\n'global' x: " << x << endl; // x = 10 (global x didn't change)
+}

### 2022-11-02 03_Code/reduction.cpp

@@ -1,0 +1,24 @@
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+using namespace std;
+
+int main() {
+  int n = 100000000;
+  // create a vector of size n, all values 1.0
+  // long double is "usually" a 128 bit float data type
+  // L indicates that 1.0 is a long double literal
+  vector<long double> vec(n, 1.0L);
+  long double sum = 0.0L;
+
+  const double start = omp_get_wtime();
+
+#pragma omp parallel for reduction(+ : sum)
+  for (int i = 0; i < n; ++i) {
+    sum += vec[i]; // sum up all values of vec
+  }
+
+  cout << fixed << "sum: " << sum << endl;
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-11-02 03_Code/segmentation_fault.cpp

@@ -1,0 +1,13 @@
+#include <iostream>
+#include <vector>
+
+using namespace std;
+
+int main() {
+  vector<int> my_vector{1, 2, 3};
+#pragma omp parallel default(none) shared(cout) private(my_vector)
+  {
+    // program crashes since my_vector is uninitialized (size = 0)
+    cout << my_vector[0]; // segmentation fault
+  }
+}

### 2022-11-02 03_Code/single.cpp

@@ -1,0 +1,15 @@
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+#pragma omp parallel
+  {
+#pragma omp single // only one thread executes the code in the single
+    {
+      cout << "from single: " << omp_get_thread_num() << "\n";
+    } // implicit barrier after single
+    cout << "after single\n";
+  };
+}

### 2022-11-09 Assignments/paper.tex
add code04

@@
 
+\section{Assignment}
+
+\subsection{How does ordered clause in OpenMP work in conjunction with a parallel loop?}
+\subsection{What is the collapse clause in OpenMP good for?}
+\subsection{Explain how reductions work internally in OpenMP}
+\subsection{What is the purpose of a barrier in parallel computing?}
+\subsection{Explain the difference between the library routines: omp\_get\_num\_threads(), omp\_get\_num\_procs() and omp\_get\_max\_threads().}
+\subsection{Clarify how the storage attributes private and firstprivate differ from each other.}
+\subsection{Write in pseudo code how the computation of pi can be parallelized with simple threads.}
+
 %
@@
 %
-% BibTeX users should specify bibliography style 'splncs04'.
+% BibTeX users should specify bibliography style 'splncs04'
 % References will then be sorted and formatted in the correct style.
@@
 
-%\bibitem{ref_book1}
-%Arpaci-Dusseau, Remzi H and Arpaci-Dusseau, Andrea C.: Operating systems: Three easy pieces. Arpaci-Dusseau Books LLCr,
- %(2018)
-
 \bibitem{ref_url1}
@@
 
-\bibitem{ref_url2}
-Armv8-A Address Translation pdf, \url{https://developer.arm.com/documentation/100940/0101/}.
-
 \end{thebibliography}

### 2022-11-09 04_Code/if_clause.cpp

@@ -1,0 +1,10 @@
+#include <iostream>
+#include <omp.h>
+using namespace std;
+
+int main() {
+  volatile int problem_size = 100;
+  // if the problem size is small, don't execute in parallel
+#pragma omp parallel if (problem_size > 1000)
+  cout << "Hello from thread " << omp_get_thread_num() << endl;
+}

### 2022-11-09 04_Code/merge_multi_threaded.cpp

@@ -1,0 +1,121 @@
+#include <algorithm>
+#include <iostream>
+#include <omp.h>
+#include <parallel/algorithm>
+#include <random>
+using namespace std;
+
+vector<int> get_random_int_vector(int n) {
+  default_random_engine re{random_device{}()};
+  uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
+  vector<int> v(n);
+  for (auto &num : v) {
+    num = next_rand(re);
+  }
+  return v;
+}
+
+/*
+ * algorithm for merge as described in:
+ *
+ * "Introduction to Algorithms" 3rd edition, p. 799
+ */
+template <class _Type, class _Compare>
+inline int64_t my_binary_search(_Type value, const _Type *a, int left,
+                                int right, _Compare __comp) {
+  int64_t low = left;
+  int64_t high = std::max(left, right + 1);
+  while (low < high) {
+    long mid = (low + high) / 2;
+    if (__comp(value, a[mid]))
+      high = mid;
+    else
+      low = mid + 1;
+  }
+  return high;
+}
+
+template <class _Type, class _Compare>
+inline void merge_ptr(const _Type *a_start, const _Type *a_end,
+                      const _Type *b_start, const _Type *b_end, _Type *dst,
+                      _Compare __comp) {
+  while (a_start < a_end && b_start < b_end) {
+    if (__comp(*a_start, *b_start))
+      *dst++ = *a_start++;
+    else
+      *dst++ = *b_start++;
+  }
+  while (a_start < a_end)
+    *dst++ = *a_start++;
+  while (b_start < b_end)
+    *dst++ = *b_start++;
+}
+
+template <class _Type, class _Compare>
+inline void merge_dac(_Type *t1, _Type *t2, int64_t p1, int64_t r1, int64_t p2,
+                      int64_t r2, _Type *a, int64_t p3, _Compare __comp) {
+  int64_t length1 = r1 - p1 + 1;
+  int64_t length2 = r2 - p2 + 1;
+  if (length1 < length2) {
+    std::swap(p1, p2);
+    std::swap(r1, r2);
+    std::swap(length1, length2);
+    std::swap(t1, t2);
+  }
+  if (length1 == 0)
+    return;
+  if ((length1 + length2) <= 8192 * 8) {
[…]

### 2022-11-09 04_Code/merge_single_threaded.cpp

@@ -1,0 +1,121 @@
+#include <algorithm>
+#include <iostream>
+#include <omp.h>
+#include <parallel/algorithm>
+#include <random>
+using namespace std;
+
+vector<int> get_random_int_vector(int n) {
+  default_random_engine re{random_device{}()};
+  uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
+  vector<int> v(n);
+  for (auto &num : v) {
+    num = next_rand(re);
+  }
+  return v;
+}
+
+/*
+ * algorithm for merge as described in:
+ *
+ * "Introduction to Algorithms" 3rd edition, p. 799
+ */
+template <class _Type, class _Compare>
+inline int64_t my_binary_search(_Type value, const _Type *a, int left,
+                                int right, _Compare __comp) {
+  int64_t low = left;
+  int64_t high = std::max(left, right + 1);
+  while (low < high) {
+    long mid = (low + high) / 2;
+    if (__comp(value, a[mid]))
+      high = mid;
+    else
+      low = mid + 1;
+  }
+  return high;
+}
+
+template <class _Type, class _Compare>
+inline void merge_ptr(const _Type *a_start, const _Type *a_end,
+                      const _Type *b_start, const _Type *b_end, _Type *dst,
+                      _Compare __comp) {
+  while (a_start < a_end && b_start < b_end) {
+    if (__comp(*a_start, *b_start))
+      *dst++ = *a_start++;
+    else
+      *dst++ = *b_start++;
+  }
+  while (a_start < a_end)
+    *dst++ = *a_start++;
+  while (b_start < b_end)
+    *dst++ = *b_start++;
+}
+
+template <class _Type, class _Compare>
+inline void merge_dac(_Type *t1, _Type *t2, int64_t p1, int64_t r1, int64_t p2,
+                      int64_t r2, _Type *a, int64_t p3, _Compare __comp) {
+  int64_t length1 = r1 - p1 + 1;
+  int64_t length2 = r2 - p2 + 1;
+  if (length1 < length2) {
+    std::swap(p1, p2);
+    std::swap(r1, r2);
+    std::swap(length1, length2);
+    std::swap(t1, t2);
+  }
+  if (length1 == 0)
+    return;
+  if ((length1 + length2) <= 8192 * 8) {
[…]

### 2022-11-09 04_Code/merge_sort.cpp

@@ -1,0 +1,100 @@
+#include <algorithm>
+#include <cstring>
+#include <iostream>
+#include <omp.h>
+#include <random>
+using namespace std;
+
+vector<int> get_random_int_vector(int n) {
+  default_random_engine re{random_device{}()};
+  uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
+  vector<int> v(n);
+  for (auto &num : v) {
+    num = next_rand(re);
+  }
+  return v;
+}
+
+// https://www.geeksforgeeks.org/insertion-sort/
+void insertion_sort(int *arr, int n) {
+  int i, key, j;
+  for (i = 1; i < n; i++) {
+    key = arr[i];
+    j = i - 1;
+
+    /* move elements of arr[0..i-1], that are
+    greater than key, to one position ahead
+    of their current position */
+    while (j >= 0 && arr[j] > key) {
+      arr[j + 1] = arr[j];
+      j = j - 1;
+    }
+    arr[j + 1] = key;
+  }
+}
+
+inline void merge(const int *__restrict__ a, const int *__restrict__ b,
+                  int *__restrict__ c, const int a_size, const int b_size,
+                  const int c_size) {
+  int idx_a = 0;
+  int idx_b = 0;
+  for (int i = 0; i < c_size; ++i) {
+    if (idx_a == a_size) {
+      c[i] = b[idx_b++];
+    } else if (idx_b == b_size) {
+      c[i] = a[idx_a++];
+    } else {
+      c[i] = (a[idx_a] < b[idx_b]) ? a[idx_a++] : b[idx_b++];
+    }
+  }
+}
+
+void merge_sort(int *arr, int n) {
+  if (n > 1) {
+    if (n < 32) { // insertion sort for n smaller than 32
+      insertion_sort(arr, n);
+      return;
+    }
+    const int size_a = n / 2;
+    const int size_b = n - size_a;
+#pragma omp task if (size_a > 10000) // make task if size_a > 10000, one task is enough
+    merge_sort(arr, size_a); // parallel recursive call
+    merge_sort(arr + size_a, size_b); // recursive call
+#pragma omp taskwait // wait until both subarrays are sorted
+    if (n < 8192) { // allocate array on the stack for small n
+      // sometimes called _alloca() or _malloca(), or "int c[n]"
+      int* c = (int*) alloca(n * sizeof(int));
[…]

### 2022-11-09 04_Code/merge_sort_no_copying.cpp

@@ -1,0 +1,102 @@
+#include <algorithm>
+#include <iostream>
+#include <omp.h>
+#include <random>
+using namespace std;
+
+vector<int> get_random_int_vector(int n) {
+  default_random_engine re{random_device{}()};
+  uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
+  vector<int> v(n);
+  for (auto &num : v) {
+    num = next_rand(re);
+  }
+  return v;
+}
+
+// https://www.geeksforgeeks.org/insertion-sort/
+void insertion_sort(int *arr, int n) {
+  int i, key, j;
+  for (i = 1; i < n; i++) {
+    key = arr[i];
+    j = i - 1;
+
+    /* move elements of arr[0..i-1], that are
+    greater than key, to one position ahead
+    of their current position */
+    while (j >= 0 && arr[j] > key) {
+      arr[j + 1] = arr[j];
+      j = j - 1;
+    }
+    arr[j + 1] = key;
+  }
+}
+
+inline void merge(const int *__restrict__ a, const int *__restrict__ b,
+                  int *__restrict__ c, const int a_size, const int b_size,
+                  const int c_size) {
+  int idx_a = 0;
+  int idx_b = 0;
+  for (int i = 0; i < c_size; ++i) {
+    if (idx_a == a_size) {
+      c[i] = b[idx_b++];
+    } else if (idx_b == b_size) {
+      c[i] = a[idx_a++];
+    } else {
+      c[i] = (a[idx_a] < b[idx_b]) ? a[idx_a++] : b[idx_b++];
+    }
+  }
+}
+
+void merge_sort(int *arr, int *buffer, int n); // forward declaration of merge_sort
+
+void merge_sort_helper(int *arr, int *buffer, int n) { // helps to implement no copying strategy
+  const int size_a = n / 2;
+  const int size_b = n - size_a;
+#pragma omp task final(size_a < 10000) // stop creating tasks if size_a < 10000
+  merge_sort(arr, buffer, size_a); // parallel recursive call to merge sort
+  merge_sort(arr + size_a, buffer + size_a, size_b); // recursive call to merge sort
+#pragma omp taskwait // wait until both subarrays are sorted
+  merge(arr, arr + size_a, buffer, size_a, size_b, n); // merge from arr into buffer
+}
+
+void merge_sort(int *arr, int *buffer, int n) {
+  if (n > 1) { // only this function can end a recursion
[…]

### 2022-11-09 04_Code/nested_parallelism.cpp

@@ -1,0 +1,22 @@
+#include <iostream>
+#include <omp.h>
+using namespace std;
+
+int main() {
+  omp_set_nested(1); // enable nested parallelism
+  int counter = 0; // counter for total threads executed
+#pragma omp parallel num_threads(4) // start outer parallel region
+  {
+    // each thread creates two new threads
+#pragma omp parallel num_threads(2) // start inner parallel region
+    {
+#pragma omp critical
+      { // thread id in inner parallel region has nothing
+        // to do with the thread id of the outer parallel region
+        cout << omp_get_thread_num() << endl; // prints 0 or 1
+        ++counter;
+      }
+    } // end inner parallel region
+  } // end outer parallel region
+  cout << "counter: " << counter << endl; // prints 8 since 4 * 2 = 8
+}

### 2022-11-09 04_Code/slow_merge_sort.cpp

@@ -1,0 +1,65 @@
+#include <algorithm>
+#include <cstring>
+#include <iostream>
+#include <omp.h>
+#include <random>
+using namespace std;
+
+vector<int> get_random_int_vector(int n) {
+  default_random_engine re{random_device{}()};
+  uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
+  vector<int> v(n);
+  for (auto &num : v) {
+    num = next_rand(re);
+  }
+  return v;
+}
+
+inline void merge(const int *__restrict__ a, const int *__restrict__ b,
+                  int *__restrict__ c, const int a_size, const int b_size,
+                  const int c_size) {
+  int idx_a = 0;
+  int idx_b = 0;
+  for (int i = 0; i < c_size; ++i) {
+    if (idx_a == a_size) {
+      c[i] = b[idx_b++];
+    } else if (idx_b == b_size) {
+      c[i] = a[idx_a++];
+    } else {
+      c[i] = (a[idx_a] < b[idx_b]) ? a[idx_a++] : b[idx_b++];
+    }
+  }
+}
+
+void merge_sort_naive(int *arr, int n) { // slow merge sort
+  if (n > 1) { // TODO: use insertion sort for small n
+    const int size_a = n / 2;
+    const int size_b = n - size_a;
+    // TODO: make next recursive call a task
+    merge_sort_naive(arr, size_a); // recursive call
+    merge_sort_naive(arr + size_a, size_b); // recursive call
+    // TODO: here should be a taskwait
+    int *c = new int[n]; // TODO: avoid using heap for small n
+    merge(arr, arr + size_a, c, size_a, size_b, n);
+    memcpy(arr, c, sizeof(int) * n);
+    delete[](c);
+  }
+}
+
+int main(int argc, char *argv[]) {
+  const int n = 100000000;
+  vector<int> v = get_random_int_vector(n);
+  vector<int> v_copy = v;
+
+  double start = omp_get_wtime();
+  merge_sort_naive(v.data(), n);
+  cout << "naive: " << omp_get_wtime() - start << " seconds" << endl;
+
+  start = omp_get_wtime();
+  sort(begin(v_copy), end(v_copy));
+  cout << "std::sort: " << omp_get_wtime() - start << " seconds" << endl;
+
+  if (v != v_copy) {
+    cout << "sort implementation is buggy\n";
+  }
+}

### 2022-11-09 04_Code/storage_attributes_task.cpp

@@ -1,0 +1,24 @@
+#include <iostream>
+using namespace std;
+
+int main() {
+  int x = 1; // x is shared in a task per default -> outside parallel region
+#pragma omp parallel num_threads(4) // begin parallel region
+  {
+    int y = 7; // y is firstprivate in a task per default -> inside parallel region
+#pragma omp single // the code in a single is executed only once
+    {
+      cout << "x: ";
+      for (int i = 0; i < 8; ++i) { // create 8 tasks
+#pragma omp task firstprivate(x) shared(y) // overwrite storage attributes
+        {
+          cout << ++x; // prints always 2 because x is firstprivate in each task
+          ++y; // this y comes from the thread that creates the single
+          // but y is updated from multiple threads -> possible race condition
+        }
+      }
+    } // implicit barrier of single
+#pragma omp critical
+    cout << "\ny: " << y; // prints 7 7 7 and (very likely) one 15
+  } // end parallel region
+}

### 2022-11-09 04_Code/task_example.cpp

@@ -1,0 +1,21 @@
+#include <iostream>
+
+using namespace std;
+
+// There are two tasks. One prints the word "race ".
+// The other one prints "car ". The order in which tasks are
+// executed is runtime dependent and may vary across multiple runs.
+int main() {
+#pragma omp parallel
+  { // we need the single construct to guarantee that
+    // each task is generated only once
+#pragma omp single // one thread generates the tasks
+    {
+      // Each #pragma omp task directive defines a task
+#pragma omp task // define task and put it in the task queue
+      { cout << "race "; } // task 1
+#pragma omp task // define task and put it in the task queue
+      { cout << "car "; } // task 2
+    } // end of single region (implicit barrier)
+  } // end of parallel region
+}

### 2022-11-09 04_Code/taskwait.cpp

@@ -1,0 +1,21 @@
+#include <iostream>
+
+using namespace std;
+
+int main() {
+#pragma omp parallel
+  { // we need the single construct to guarantee that
+    // each task is generated only once
+#pragma omp single // one thread generates the tasks
+    {
+      cout << "A "; // printed before tasks are generated
+      // Each #pragma omp task directive defines a task
+#pragma omp task // define task and put it in the task queue
+      { cout << "race "; } // task 1
+#pragma omp task // define task and put it in the task queue
+      { cout << "car "; } // task 2
+#pragma omp taskwait // wait here until the tasks are completed
+      cout << "is fun to watch.\n";
+    } // end of single region
+  } // end of parallel region
+}

### 2022-11-09 Assignments/paper.tex
Assignment 3

@@
 \documentclass[runningheads]{llncs}
-
 \usepackage[T1]{fontenc}
 \usepackage{graphicx}
+\usepackage{listings}
+
+\lstset{
+	language=C++,
+    %basicstyle=\ttfamily,
+    %keywordstyle=\color{blue}\ttfamily,
+    %stringstyle=\color{red}\ttfamily,
+    %commentstyle=\color{green}\ttfamily,
+    %morecomment=[l][\color{magenta}]{\#}
+}
 
@@
 
-% \begin{abstract}
-%
-% \end{abstract}
-
+% (1)
 \section{Assignment}
@@
 
+% (2)
 \section{Assignment}
@@
 
+% (3)
 \section{Assignment}
@@
 \subsection{How does ordered clause in OpenMP work in conjunction with a parallel loop?}
+An ordered clause is used within a parallel region. It opens a region where the execution is sequential, in order to prevent race conditions.
+
 \subsection{What is the collapse clause in OpenMP good for?}
+The collapse clause is used to parallelize for loops.
+It allows to specify how many loops are collapsed into one.
+The collapse clause is useful for balancing work and parallelize multiple loops to boost performance.
+When using the collapse clause a single loop is formed. The length of the new loop is equal to the multiplication of the length of each loop that has to be collapsed into one.
+
 \subsection{Explain how reductions work internally in OpenMP}
+A reduction is an operation form OpenMP that is used in parallel computing. 
+It reduces the calues into a single result.
+The syntax of the operation is: \lstinline{reduction(op:list)}. op refers to the operation, which can be +, -, *, and so on. list refers to a list of variables that is reduced.
+For example: A reduction can be used within a parallel for loop: 
+\lstinline{#pragma omp parallel for reduction(+ : sum) } to compute the sum of the values 1, 2, 3.
+Every thread within the parallel region gets a local sum variable, where the sum is computed.
+The result will be written to the global sum variable declared outside of the parallel region.
+
 \subsection{What is the purpose of a barrier in parallel computing?}
+A barrier can be set within a parallel region.
+Setting a barrier means that the threads will execute the code until that specific barrier.
+There they wait until all threads have reached the barrier and only then continue with execution.
+Barriers are used for synchronization in parallel regions, where they enforce a specific execution order. \\
+Within OpenMP there are also implicit barriers.
+The most general barrier is the omp parallel region itself.
+The region makes sure that execution is parallel within that region.
+After the code from that region is executed in parallel execution will continue sequentially.
+
 \subsection{Explain the difference between the library routines: omp\_get\_num\_threads(), omp\_get\_num\_procs() and omp\_get\_max\_threads().}
+\begin{itemize}
+	\item omp\_get\_num\_threads() - 
+	The routine is called inside a parallel region. It is used to get the number of threads used within a parallel region.
+	\item omp\_get\_num\_procs() - 
+	The routine is used to get the number of logical cored.
+It can be used to define the number of threads reasonable to use in a parallel region.
+Per default the number of threads is equal to the number of logical cores.
+	\item omp\_get\_max\_threads() - 
+	The routine is called outside a parallel region and returns the number of threads supported in a parallel region.
+When the value is changed with omp\_set\_num\_threads() within a parallel region, the method returns the set value.
+\end{itemize}
+
 \subsection{Clarify how the storage attributes private and firstprivate differ from each other.}
+Storage attributes are used to specify the usage of variables within a parallel region. It is used for variables declared outside a parallel region.
+There are three different storage attributes: shared, firstprivate and private.
+
+Private attributes make sure that each thread gets an uninitialized copy of the variable.
+Firstprivate attributes make sure that eacht thrad gets an identical initialized copy of the variable. The variable is a local variable for each thread.
+When using private or firstprivate storage attributes the value of the global variable remains unchanged.
+
 \subsection{Write in pseudo code how the computation of pi can be parallelized with simple threads.}
+\begin{lstlisting} [language=TeX]
+	initialize num_steps to 100000000
+	initialize width to 1.0
+	initialize sum to 0.0
+
+	do in parallel:
+	  initialize sum_local to 0.0
+	  for i in {0, num_steps} do
+	    set x to ((i + 0.5) * width)
+		add sum_local to (1.0 / (1.0 + x * x))
+
+		do sequentially:
+		  add sum sum_local
+
+	initialize pi to sum * 4 * width
+	return pi
+
+\end{lstlisting}
+
+% (4)
+\section{Assignment}
 

### 2022-11-16 Assignments/paper.tex
update

@@
 
+\subsection{Explain how divide and conquer algorithms can be parallelized with tasks in OpenMP.}
+\subsection{Describe some ways to speed up merge sort.}
+\subsection{What is the idea behind multithreaded merging?}
+\subsection{Read What every systems programmer should know about concurrency. https://assets.bitbashing.io/papers/concurrency- primer.pdf
+Discuss two things you find particularly interesting.}
+
 %

### 2022-11-16 05_Code/lower_fast.cpp

@@ -1,0 +1,62 @@
+#include <algorithm>
+#include <cstring>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+// fill char array of size n with random alphanumeric strings
+void gen_random(char *s, const int64_t n) {
+  static const char alphanum[] = "0123456789"
+                                 "ABCDEFGHIJKLMNOPQRSTUVWXYZ"
+                                 "abcdefghijklmnopqrstuvwxyz";
+  for (int64_t i = 0; i < n; ++i) {
+    s[i] = alphanum[rand() % (sizeof(alphanum) - 1)];
+  }
+  s[n] = 0;
+}
+
+// generates a random alphanumeric string of size n
+string generate_alpha_numeric_string(const int64_t n) {
+  string chars{};
+  chars.resize(n);
+  char *data_ptr = const_cast<char *>(chars.data());
+  gen_random(data_ptr, n);
+  return chars;
+}
+
+// convert char array to lower case
+void lower(char *s) {
+  // 'a' is greater than 'A' in ASCII
+  constexpr char addend = 'a' - 'A';
+  // compute strlen only once
+  const size_t n = strlen(s);
+  for (size_t i = 0; i < n; i++) {
+    // no conditional if
+    // now compiler is able to use vector instructions
+    bool res = 'A' <= s[i] && s[i] <= 'Z';
+    s[i] += addend * res; // or use += addend & (0 - res)
+  }
+}
+
+int main() {
+  // create 2 identical random strings of size n
+  int64_t n = 100000000;
+  string chars = generate_alpha_numeric_string(n);
+  char *data_ptr = const_cast<char *>(chars.data());
+  string chars2 = chars; // copy chars into chars2
+
+  // custom implementation of converting a string to lower case
+  double start = omp_get_wtime();
+  lower(data_ptr);
+  cout << "custom: " << omp_get_wtime() - start << " seconds" << endl;
+
+  // the C++ way of converting a string to lower case
+  start = omp_get_wtime();
+  std::transform(chars2.begin(), chars2.end(), chars2.begin(),
+                 [](char c) { return std::tolower(c); });
+  cout << "C++: " << omp_get_wtime() - start << " seconds" << endl;
+
+  // test if strings are the same
+  cout << boolalpha << "same: " << ((chars == chars2)) << endl;
+}

### 2022-11-16 05_Code/lower_slow.cpp

@@ -1,0 +1,59 @@
+#include <algorithm>
+#include <cstring>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+// fill char array of size n with random alphanumeric strings
+void gen_random(char *s, const int64_t n) {
+  static const char alphanum[] = "0123456789"
+                                 "ABCDEFGHIJKLMNOPQRSTUVWXYZ"
+                                 "abcdefghijklmnopqrstuvwxyz";
+  for (int64_t i = 0; i < n; ++i) {
+    s[i] = alphanum[rand() % (sizeof(alphanum) - 1)];
+  }
+  s[n] = 0;
+}
+
+// generates a random alphanumeric string of size n
+string generate_alpha_numeric_string(const int64_t n) {
+  string chars{};
+  chars.resize(n);
+  char *data_ptr = const_cast<char *>(chars.data());
+  gen_random(data_ptr, n);
+  return chars;
+}
+
+// convert char array to lower case
+void lower(char *s) {
+  // 'a' is greater than 'A' in ASCII
+  constexpr char addend = 'a' - 'A';
+  for (size_t i = 0; i < strlen(s); i++) {
+    if ('A' <= s[i] && s[i] <= 'Z') {
+      s[i] += addend;
+    }
+  }
+}
+
+int main() {
+  // create 2 identical random strings of size n
+  int64_t n = 100000;
+  string chars = generate_alpha_numeric_string(n);
+  char *data_ptr = const_cast<char *>(chars.data());
+  string chars2 = chars; // copy chars into chars2
+
+  // custom implementation of converting a string to lower case
+  double start = omp_get_wtime();
+  lower(data_ptr);
+  cout << "custom: " << omp_get_wtime() - start << " seconds" << endl;
+
+  // the C++ way of converting a string to lower case
+  start = omp_get_wtime();
+  std::transform(chars2.begin(), chars2.end(), chars2.begin(),
+            [](char c) { return std::tolower(c); });
+  cout << "C++: " << omp_get_wtime() - start << " seconds" << endl;
+
+  // test if strings are the same
+  cout << boolalpha << "same: " << ((chars == chars2)) << endl;
+}

### 2022-11-19 Assignments/paper.tex
add 04_Code & Exam Assignments

--- 2022-11-09 Assignments/paper.tex
@@
 
+\subsection{Explain how divide and conquer algorithms can be parallelized with tasks in OpenMP.}
+Divide and conquer algorithms recursively split down a problem until the pieces can be directly solved.
+Such algorithms can be paralleized in OpenMP using the constructs taks and taskwait for the recursive call.
+OpenMP tasks are independent work blocks. A queueing system dynamically handles the assignment of threads to the tasks. 
+Each threads picks up a task from the queue and executes it. This goes on until the queue is empty. 
+Tasks are defined within a parallel region and within a single construct, to make sure that one thread creates the tasks.
+Taskwait is a construct that waits until all tasks are executed.
+
+\subsection{Describe some ways to speed up merge sort.}
+\begin{itemize}
+	\item Tasks - With task and taskwait
+	\item Avoid copying - to prevent that additional memory is created recursively and make sure that initially memory is instantiated and reused.
+	\item if Clause - task will only be created, if a certain condition is true
+	\item final Clause - tasks won't be created anymore, if a condition is true
+\end{itemize}
+
+\subsection{What is the idea behind multithreaded merging?}
+The goal is to sort and merge two arrays using a divide and conquer algorithm. To do that we look at the median of the
+array with more elements. Then we look at the other array and compare the elements smaller than the median and merge these
+with the respective elements from the other array. That way merge is executed in parallel.
+
+\subsection{Read What every systems programmer should know about concurrency. https://assets.bitbashing.io/papers/concurrency- primer.pdf
+Discuss two things you find particularly interesting.}
+The order of written code can be changed as compiler try to optimize it and rewrite code to run faster on the targeted hardware.
+RAM hasn’t speeded up the way CPU processors have. This creates a widening gap between the instruction fetch and the time needed to retrieve data from memory.
+To prevent reordering we can use atomic types in C or C++.
+
+Atomicity means that something can’t be divided into smaller pieces. Threads need to use atomic reads and writes to share data. 
+If they are not atomic they are so-called torn reads and writes.
+To ensure atomicity one need to make sure that variables used for thread synchronization aren’t larger than the CPU word size.
 %

### 2022-11-19 04_Code/if_clause.cpp

@@ -2,8 +2,10 @@
 #include <omp.h>
 using namespace std;
 
-int main() {
-  volatile int problem_size = 100;
+int main()
+{
+  volatile int problem_size = 1001;
+
   // if the problem size is small, don't execute in parallel
 #pragma omp parallel if (problem_size > 1000)
   cout << "Hello from thread " << omp_get_thread_num() << endl;

### 2022-11-19 04_Code/merge_single_threaded.cpp

@@ -5,11 +5,13 @@
 #include <random>
 using namespace std;
 
-vector<int> get_random_int_vector(int n) {
+vector<int> get_random_int_vector(int n)
+{
   default_random_engine re{random_device{}()};
   uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
   vector<int> v(n);
-  for (auto &num : v) {
+  for (auto &num : v)
+  {
     num = next_rand(re);
   }
   return v;
@@ -22,10 +24,12 @@ vector<int> get_random_int_vector(int n) {
  */
 template <class _Type, class _Compare>
 inline int64_t my_binary_search(_Type value, const _Type *a, int left,
-                                int right, _Compare __comp) {
+                                int right, _Compare __comp)
+{
   int64_t low = left;
   int64_t high = std::max(left, right + 1);
-  while (low < high) {
+  while (low < high)
+  {
     long mid = (low + high) / 2;
     if (__comp(value, a[mid]))
       high = mid;
@@ -38,8 +42,10 @@ inline int64_t my_binary_search(_Type value, const _Type *a, int left,
 template <class _Type, class _Compare>
 inline void merge_ptr(const _Type *a_start, const _Type *a_end,
                       const _Type *b_start, const _Type *b_end, _Type *dst,
-                      _Compare __comp) {
-  while (a_start < a_end && b_start < b_end) {
+                      _Compare __comp)
+{
+  while (a_start < a_end && b_start < b_end)
+  {
     if (__comp(*a_start, *b_start))
       *dst++ = *a_start++;
     else
@@ -53,10 +59,12 @@ inline void merge_ptr(const _Type *a_start, const _Type *a_end,
 
 template <class _Type, class _Compare>
 inline void merge_dac(_Type *t1, _Type *t2, int64_t p1, int64_t r1, int64_t p2,
-                      int64_t r2, _Type *a, int64_t p3, _Compare __comp) {
+                      int64_t r2, _Type *a, int64_t p3, _Compare __comp)
+{
   int64_t length1 = r1 - p1 + 1;
   int64_t length2 = r2 - p2 + 1;
-  if (length1 < length2) {
+  if (length1 < length2)
+  {
     std::swap(p1, p2);
     std::swap(r1, r2);
[…]

### 2022-11-19 04_Code/merge_sort.cpp

@@ -5,27 +5,32 @@
 #include <random>
 using namespace std;
 
-vector<int> get_random_int_vector(int n) {
+vector<int> get_random_int_vector(int n)
+{
   default_random_engine re{random_device{}()};
   uniform_int_distribution<int> next_rand{INT32_MIN, INT32_MAX};
   vector<int> v(n);
-  for (auto &num : v) {
+  for (auto &num : v)
+  {
     num = next_rand(re);
   }
   return v;
 }
 
 // https://www.geeksforgeeks.org/insertion-sort/
-void insertion_sort(int *arr, int n) {
+void insertion_sort(int *arr, int n)
+{
   int i, key, j;
-  for (i = 1; i < n; i++) {
+  for (i = 1; i < n; i++)
+  {
     key = arr[i];
     j = i - 1;
 
     /* move elements of arr[0..i-1], that are
     greater than key, to one position ahead
     of their current position */
-    while (j >= 0 && arr[j] > key) {
+    while (j >= 0 && arr[j] > key)
+    {
       arr[j + 1] = arr[j];
       j = j - 1;
     }
@@ -35,35 +40,47 @@ void insertion_sort(int *arr, int n) {
 
 inline void merge(const int *__restrict__ a, const int *__restrict__ b,
                   int *__restrict__ c, const int a_size, const int b_size,
-                  const int c_size) {
+                  const int c_size)
+{
   int idx_a = 0;
   int idx_b = 0;
-  for (int i = 0; i < c_size; ++i) {
-    if (idx_a == a_size) {
+  for (int i = 0; i < c_size; ++i)
+  {
+    if (idx_a == a_size)
+    {
       c[i] = b[idx_b++];
-    } else if (idx_b == b_size) {
+    }
+    else if (idx_b == b_size)
+    {
       c[i] = a[idx_a++];
-    } else {
+    }
+    else
+    {
       c[i] = (a[idx_a] < b[idx_b]) ? a[idx_a++] : b[idx_b++];
     }
   }
 }
 
-void merge_sort(int *arr, int n) {
-  if (n > 1) {
-    if (n < 32) { // insertion sort for n smaller than 32
+void merge_sort(int *arr, int n)
+{
+  if (n > 1)
+  {
+    if (n < 32)
+    { // insertion sort for n smaller than 32
       insertion_sort(arr, n);
       return;
     }
     const int size_a = n / 2;
     const int size_b = n - size_a;
[…]

### 2022-11-19 04_Code/nested_parallelism.cpp

@@ -2,9 +2,10 @@
 #include <omp.h>
 using namespace std;
 
-int main() {
-  omp_set_nested(1); // enable nested parallelism
-  int counter = 0; // counter for total threads executed
+int main()
+{
+  omp_set_nested(1);                // enable nested parallelism
+  int counter = 0;                  // counter for total threads executed
 #pragma omp parallel num_threads(4) // start outer parallel region
   {
     // each thread creates two new threads
@@ -16,7 +17,10 @@ int main() {
         cout << omp_get_thread_num() << endl; // prints 0 or 1
         ++counter;
       }
-    } // end inner parallel region
-  } // end outer parallel region
+      for (volatile uint64_t i = 0; i < 1000000000; ++i)
+      {
+      }
+    }                                     // end inner parallel region
+  }                                       // end outer parallel region
   cout << "counter: " << counter << endl; // prints 8 since 4 * 2 = 8
 }

### 2022-11-19 04_Code/storage_attributes_task.cpp

@@ -1,24 +1,26 @@
 #include <iostream>
 using namespace std;
 
-int main() {
-  int x = 1; // x is shared in a task per default -> outside parallel region
+int main()
+{
+  int x = 1;                        // x is shared in a task per default -> outside parallel region
 #pragma omp parallel num_threads(4) // begin parallel region
   {
-    int y = 7; // y is firstprivate in a task per default -> inside parallel region
+    int y = 7;     // y is firstprivate in a task per default -> inside parallel region
 #pragma omp single // the code in a single is executed only once
     {
       cout << "x: ";
-      for (int i = 0; i < 8; ++i) { // create 8 tasks
+      for (int i = 0; i < 8; ++i)
+      {                                    // create 8 tasks
 #pragma omp task firstprivate(x) shared(y) // overwrite storage attributes
         {
           cout << ++x; // prints always 2 because x is firstprivate in each task
-          ++y; // this y comes from the thread that creates the single
+          ++y;         // this y comes from the thread that creates the single
           // but y is updated from multiple threads -> possible race condition
         }
       }
     } // implicit barrier of single
 #pragma omp critical
     cout << "\ny: " << y; // prints 7 7 7 and (very likely) one 15
-  } // end parallel region
+  }                       // end parallel region
 }

### 2022-11-19 04_Code/task_example.cpp

@@ -1,21 +1,32 @@
 #include <iostream>
+#include <omp.h>
 
 using namespace std;
+
+double start_time = omp_get_wtime(); // wall clock time in seconds
 
 // There are two tasks. One prints the word "race ".
 // The other one prints "car ". The order in which tasks are
 // executed is runtime dependent and may vary across multiple runs.
-int main() {
+int main()
+{
 #pragma omp parallel
-  { // we need the single construct to guarantee that
-    // each task is generated only once
+  {                // we need the single construct to guarantee that
+                   // each task is generated only once
 #pragma omp single // one thread generates the tasks
     {
       // Each #pragma omp task directive defines a task
 #pragma omp task // define task and put it in the task queue
-      { cout << "race "; } // task 1
+      {
+        cout << "race ";
+      }          // task 1
 #pragma omp task // define task and put it in the task queue
-      { cout << "car "; } // task 2
-    } // end of single region (implicit barrier)
-  } // end of parallel region
+      {
+        cout << "car ";
+      } // task 2
+    }   // end of single region (implicit barrier)
+  }     // end of parallel region
+
+  double run_time = omp_get_wtime() - start_time;
+  cout << "\ncalculated in " << run_time << " seconds\n";
 }

### 2022-11-30 04_Code/merge_single_threaded.cpp
06

@@ -84,10 +84,15 @@ inline void merge_dac(_Type *t1, _Type *t2, int64_t p1, int64_t r1, int64_t p2,
     int64_t q3 = p3 + (q1 - p1) + (q2 - p2);
     a[q3] = t1[q1];
 
-    // TODO: parallelize the first recursive call with a task
-    // #pragma omp parallel
-    merge_dac(t1, t2, p1, q1 - 1, p2, q2 - 1, a, p3, __comp);
-    merge_dac(t1, t2, q1 + 1, r1, q2, r2, a, q3 + 1, __comp);
+// TODO: parallelize the first recursive call with a task
+#pragma omp task
+    {
+      merge_dac(t1, t2, p1, q1 - 1, p2, q2 - 1, a, p3, __comp);
+    }
+#pragma omp task
+    {
+      merge_dac(t1, t2, q1 + 1, r1, q2, r2, a, q3 + 1, __comp);
+    }
   }
 }
 
@@ -95,11 +100,13 @@ template <class _Type, class _Compare>
 void parallel_merge(_Type *arr1, _Type *arr2, _Type *out,
                     int64_t size1, int64_t size2, int n_threads, _Compare __comp)
 {
-  // TODO: parallelize merge_dac, use n_threads as the amount of threads
-  // for the parallel region
-  // #pragma omp parallel num_threads(n_threads)
-  // #pragma omp single nowait
+// TODO: parallelize merge_dac, use n_threads as the amount of threads
+// for the parallel region
+#pragma omp parallel num_threads(n_threads) {
+#pragma omp single nowait {
   merge_dac(arr1, arr2, 0, size1 - 1, 0, size2 - 1, out, 0, __comp);
+}
+}
 }
 
 int main(int argc, char *argv[])

### 2022-11-30 Assignments/paper.tex
06

@@
 \subsection{Describe how parallelism differs from concurrency}
-Parallelism is a subset of concurrency. If a system is parallel it is also concurrent, but not the other way around. A system is considered parallel if two or more tasks are executed simultaneously.
+Parallelism is a subset of concurrency. If a system is parallel it is also concurrent, but not the other way around. 
+A system is considered parallel if two or more tasks are executed simultaneously.
 \\
-Concurrency only means support for two or more actions at the same time. Parallelism refers to execution of more than one action at the same time.
+Concurrency only means support for two or more actions at the same time. 
+Parallelism refers to execution of more than one action at the same time.
 
@@
 
-The processor can read data from the register file, which is within the CPU almost 100 times faster than from memory. The gap between processor and memory is continuously growing.
+The processor can read data from the register file, which is within the CPU almost 100 times faster than from memory. 
+The gap between processor and memory is continuously growing.
 
-Cache memory is used to bridge the gap.
-It is used to temporarily store information that will likely be used in the future.
+Cache memory is used to bridge the gap. It is used to temporarily store information that will likely be used in the future. 
 There exist several levels of caches -  L1, L2, L3 - which go from smallest to largest.
 
-A cache reduces the acess time to data in memory. Frequently used data and instructions are kept in the cache. Access to data kept there is faster.
+A cache reduces the acess time to data in memory. Frequently used data and instructions are kept in the cache. 
+Access to data kept there is faster.
 
 Modern desktops, servers and industrial CPUs have at least three independent caches: instruction cache, data cache and the TLB. 
-The instruction cache is used to speed up executable instruction fetches.
-The data cache speeds up data fetch and store.
+The instruction cache is used to speed up executable instruction fetches. The data cache speeds up data fetch and store. 
 It is organized into more cache levels:
@@
 	\item L1 cache - Primary cache is fast but small and usually embedded in the processor chip (CPU)
-	\item L2 cache - Secondary cache is larger and can be embedded in the CPU or on a seperate chip or coprocessor containing a high-speed alternative system bus connectin cache and CPU.
+	\item L2 cache - Secondary cache is larger and can be embedded in the CPU or on a seperate chip or coprocessor containing 
+	a high-speed alternative system bus connectin cache and CPU.
 	\item L3 cache - specialized memory to improve L1 and L2. L3 is usually double the speed of DRAM
@@
 The small caches L1 are backed up by larger and slower caches L2, L3 to address the tradeoff between cache latency and hit rate. 
-The faster cache is generally checked firt.
-If that cache misses, the next chache is checked and so on, before accessing main memory.
+The faster cache is generally checked firt. If that cache misses, the next chache is checked and so on, before accessing main memory.
 
@@
 
-The Translation lookaside buffer is used to speed up the translation from virtual to physical addresses. It is part of the memory management unit and not directly related to CPU caches.
+The Translation lookaside buffer is used to speed up the translation from virtual to physical addresses. 
+It is part of the memory management unit and not directly related to CPU caches.
 
@@
 \paragraph{Explain figure “Performance gains after Moore’s law ends”}
-The figure illustrates the “Top” and “Bottom” referring to computer performance gains. The “Bottom” refers to the miniaturization of computer components seen in the last decades. \\
+The figure illustrates the “Top” and “Bottom” referring to computer performance gains. 
+The “Bottom” refers to the miniaturization of computer components seen in the last decades. \\
 Due to physical limits the opportunities for gains at the bottom will slowly come to an end.
@@
 False sharing occurs, when threads on different processors modify variables on the same cache line. 
-A cache line is the smallest unit of memory.
-Its length depends on the underlying architecture and is typically 64 and in more recent architectures 128 bytes long.
+A cache line is the smallest unit of memory. Its length depends on the underlying architecture and is typically 64 and 
+in more recent architectures 128 bytes long.
 \\
@@
 \subsection{How do mutual exclusion constructs prevent race conditions?}
-Race conditions occur when two threads modify the same data.
-Mutual exclusions give one thread exclusive access to the data.
+Race conditions occur when two threads modify the same data. Mutual exclusions give one thread exclusive access to the data. 
 After execution exclusive access is returned and the next thread can execute.
@@
 Each thread will work on a chunk of values and then take the next chunk that hasn't been worked on by any thread.
-Static schedules perform better for balanced workloads and dynamic schedules for unbalanced workloads, in case the workload varies between different iterations of the for loop.
+Static schedules perform better for balanced workloads and dynamic schedules for unbalanced workloads, 
+in case the workload varies between different iterations of the for loop.
 The chunk size can be specifies; for static schedules it is one per default.
@@
 \subsection{What can we do if we've found a solution while running a parallel for loop in OpenMP, but still have many iterations left?}
-If we've found a solution within a parallel for loop, we can use continue. We don't continue calculating, but quickly iterate through the loop to increase performance.
+If we've found a solution within a parallel for loop, we can use continue. We don't continue calculating, 
+but quickly iterate through the loop to increase performance.
 
@@
 strd::atomic::compare\_exchange\_weak  succeeds only if the value of the variable to be updated equals the first provided argument.
-For example if for final\_solution.compare\_exchange\_weak(previous, i) final\_solution equals to previous. Previous is updated with i, if it fails.
+For example if for final\_solution.compare\_exchange\_weak(previous, i) final\_solution equals to previous. 
+Previous is updated with i, if it fails.
 
@@
 \subsection{How does ordered clause in OpenMP work in conjunction with a parallel loop?}
-An ordered clause is used within a parallel region. It opens a region where the execution is sequential, in order to prevent race conditions.
+An ordered clause is used within a parallel region. It opens a region where the execution is sequential, 
+in order to prevent race conditions.
 
 \subsection{What is the collapse clause in OpenMP good for?}
-The collapse clause is used to parallelize for loops.
-It allows to specify how many loops are collapsed into one.
+The collapse clause is used to parallelize for loops. It allows to specify how many loops are collapsed into one. 
 The collapse clause is useful for balancing work and parallelize multiple loops to boost performance.
-When using the collapse clause a single loop is formed. The length of the new loop is equal to the multiplication of the length of each loop that has to be collapsed into one.
+When using the collapse clause a single loop is formed. The length of the new loop is equal to the multiplication 
+of the length of each loop that has to be collapsed into one.
 
@@
 It reduces the calues into a single result.
-The syntax of the operation is: \lstinline{reduction(op:list)}. op refers to the operation, which can be +, -, *, and so on. list refers to a list of variables that is reduced.
+The syntax of the operation is: \lstinline{reduction(op:list)}. op refers to the operation, which can be +, -, *, and so on. 
+list refers to a list of variables that is reduced.
 For example: A reduction can be used within a parallel for loop: 
@@
 \subsection{What is the purpose of a barrier in parallel computing?}
-A barrier can be set within a parallel region.
-Setting a barrier means that the threads will execute the code until that specific barrier.
+A barrier can be set within a parallel region. Setting a barrier means that the threads will execute the code until that specific barrier. 
 There they wait until all threads have reached the barrier and only then continue with execution.
 Barriers are used for synchronization in parallel regions, where they enforce a specific execution order. \\
-Within OpenMP there are also implicit barriers.
-The most general barrier is the omp parallel region itself.
+Within OpenMP there are also implicit barriers. The most general barrier is the omp parallel region itself. 
 The region makes sure that execution is parallel within that region. 
@@
 	\item omp\_get\_num\_procs() - 
-	The routine is used to get the number of logical cored.
-It can be used to define the number of threads reasonable to use in a parallel region.
-Per default the number of threads is equal to the number of logical cores.
+	The routine is used to get the number of logical cored. It can be used to define the number of threads reasonable to use 
+	in a parallel region. Per default the number of threads is equal to the number of logical cores.
 	\item omp\_get\_max\_threads() - 
@@
 \subsection{Clarify how the storage attributes private and firstprivate differ from each other.}
-Storage attributes are used to specify the usage of variables within a parallel region. It is used for variables declared outside a parallel region.
+Storage attributes are used to specify the usage of variables within a parallel region. 
+It is used for variables declared outside a parallel region.
 There are three different storage attributes: shared, firstprivate and private.
@@
 Private attributes make sure that each thread gets an uninitialized copy of the variable.
-Firstprivate attributes make sure that eacht thrad gets an identical initialized copy of the variable. The variable is a local variable for each thread.
+Firstprivate attributes make sure that eacht thrad gets an identical initialized copy of the variable. 
+The variable is a local variable for each thread.
 When using private or firstprivate storage attributes the value of the global variable remains unchanged.
@@
 	\item Tasks - With task and taskwait
-	\item Avoid copying - to prevent that additional memory is created recursively and make sure that initially memory is instantiated and reused.
+	\item Avoid copying - to prevent that additional memory is created recursively and make sure that initially memory 
+	is instantiated and reused.
 	\item if Clause - task will only be created, if a certain condition is true
@@
 The order of written code can be changed as compiler try to optimize it and rewrite code to run faster on the targeted hardware.
-RAM hasn’t speeded up the way CPU processors have. This creates a widening gap between the instruction fetch and the time needed to retrieve data from memory.
+RAM hasn’t speeded up the way CPU processors have. This creates a widening gap between the instruction fetch and the time 
+needed to retrieve data from memory.
 To prevent reordering we can use atomic types in C or C++.
@@
 To ensure atomicity one need to make sure that variables used for thread synchronization aren’t larger than the CPU word size.
+
+% (5)
+\section{Assignment}
+
+\subsection{What is CMake?}
+CMake is a script language used to create build files that can be executed across different platforms.
+The build files are executed using a compiler. CMake provides tools to build, test and package software.
+
+\subsection{What role do targets play in CMake?}
+Targets define what is build. They are executables and libraries.
+Targets have constructors to build executables, libraries or tests.
+Targets can be thought of as objects with different properties. To change these properties we can use member functions.
+Within member functions the properties can be specified. Properties can flags, directories or linked libraries.
+
+\subsection{How would you proceed to optimize code?}
+The most important thing is to try solving the problem. Once that is done and my code works, I would look at the performance.
+If performance can be improved, I would look at ways to do so. I would try parallelizing sections of code, try finding ways to improve the algorithm.
+I would continue doing so until the performance is good.
+
+% (6)
+\section{Assignment}
+
+\subsection{Name some characteristics of the instruction sets: SSE, AVX(2), AVX-512}
+SSE, AVX2, AVX-512 refer to instruction sets on Intel CPUs. They differ in the supported vector length and the number of registers.
+
+SSE: 128-bit vector length, 1999 - 2009 years of launch, 16 registers
+AVX2: 256-bit vector length, 2011 / 2013 years of launch, 16 registers
+AVX-512: 512-bit vector length, 2017 year of launch, 32 registers
+
+CPUs for high-performance computing support AVX-512 instruction set, wehere as modern mainstream CPUs support AVX2 at most.
+
+\subsection{How can memory aliasing affect performance?}
+Memory aliasing occurs when two pointers point to the same memory location.
+The compiler doesn't know if two pointers might point to the same address. 
+To ensure safe optimizations, the compiler always assumes pointer aliasing. 
+Therefore, we can give the compiler hints, to ensure it that no memory aliasing was made.
+This can be done using the restrict keyword (\*\_\_restrict\_\_ for gcc). Memory aliasing is important for pointers as well as references.
+
+\subsection{What are the advantages of unit stride (stride-1) memory access compared to accessing memory with larger strides (for example stride-8)?}
+stride-1 - access sequential elements in memory, where each fields are equally distant. The distance in this case is called stride.
+The difference between stride-1 and larger strides stride-n is that every nth element is accessed.
+
+Stride-1 is faster than larger strides. There is a maximum bandwith that can be loaded from memory. 
+Using larger strides leads to loading a higher bandwidth without fully using it.
+
+We can check this by looking at the summation of two arrays. 
+Summing up every 8th element is more or less as performant as summing up every element of the array.
+
+\subsection{When would you prefer arranging records in memory as a Structure of Arrays?}
+There are two ways to arrange records in memory: as Array of Structures or Structure of Arrays.
+Structure of Arrays: one struct with multiple arrays.
+Array of Structures: one array with the elements for each entry.
+
+For example when we want to save x,y,z coordinates. When using a Structure of Arrays we have one struct and three arrays for each coordinate.
+If we use an Array of Structures we have a struct array with three entries. Each entry stores the values of the x, y, z coordinates.
+
+A Structure of Arrays is good for vectorization. In our example with the coordinates, we can access all the entries of a specific coordinate
+through the respective array.
+With a SoA the arrays are kept seperate for each structure field. Memory access is contiguous if we perform vectorization of the structure instances.
+When performing operations on one array of the structure, SoA leads better bandwidth usage.
+
 %

### 2022-11-30 template_paper/paper.tex
06

\documentclass[sigconf]{acmart}
\usepackage{booktabs}

%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% These commands are for a PROCEEDINGS abstract or paper.
\settopmatter{printacmref=false} % Removes citation information below abstract
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with conference information in

\acmConference[AEPRO 2023]{AEPRO 2023: Algorithm Engineering Projects}{March 1}{Jena, Germany}

% convert text to title case
% http://individed.com/code/to-title-case/

% that helps you to formulate your sentences
% https://www.deepl.com/translator

\begin{document}

%%
%% The "title" command has an optional parameter,
%% allowing the author to define a "short title" to be used in page headers.
\title[Fast Entity Resolution With Mock Labels and Sorted Integer Sets]{Fast Entity Resolution With Mock Labels and Sorted Integer Sets\\\large Algorithm Engineering 2023 Project Paper}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.

\author{Sheela Orgler}
\affiliation{%
  \institution{Friedrich Schiller University Jena}
  \country{Germany}}
\email{sheela.orgler@uni-jena.de}

%%\author{Erika Mustermann}
%%\affiliation{%
  % \institution{Friedrich Schiller University Jena}
  % \country{Germany}}
% \email{erika.mustermann@uni-jena.de}

%% The abstract is a short summary of the work to be presented in the article.
\begin{abstract}

The five-finger pattern \cite{macgilchrist2014}:
\begin{enumerate}
\item \textbf{Topic and background:} What topic does the paper deal with? What is the point of departure for your research? Why are you studying this now?
\item \textbf{Focus:} What is your research question? What are you studying precisely?
\item \textbf{Method:} What did you do?
\item \textbf{Key findings:} What did you discover?
\item \textbf{Conclusions or implications:} What do these findings mean? What broader issues do they speak to?
\end{enumerate}

\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{entity resolution, data cleansing, programming contest}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

\let\thefootnote\relax\footnotetext{AEPRO 2023, March 1, Jena, Germany. Copyright \copyright 2023 for this paper by its authors. Use permitted under Creative Commons License Attribution 4.0 International (CC BY 4.0).}

\section{Introduction}

\subsection{Background}

\subsection{Related Work}
%% Paper: Modified Adaptive Thresholding Using Integral Image

\subsection{Our Contributions}

\subsection{Outline}

\section{The Algorithm}

\subsection{Internal Representation of Mock Labels}
\label{sub:sec:internal}

In Figure~\ref{fig:integer:sets} we convert the mock labels to sorted integer sets.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=\linewidth]{./graphics/integer_sets.pdf}
  \caption{Conversion of mock camera labels to sorted integer sets. 
We map each unique token (key) in camera labels to a unique value. 
Based on these key-value-mappings, we convert camera labels to sorted integer sets. 
A camera can have different names in different countries. Therefore, repeating IDs reference the same cameras (see, for example, ID=3).} 
  \label{fig:integer:sets}
\end{figure}

\subsection{Efficient Preprocessing of Input Data}
\label{sub:sec:preprocessing}

The following findings are important to speed up preprocessing of the input data:

\begin{itemize}
\item Reading many small files concurrently, with multiple threads (compared to a single thread), takes advantage of the internal parallelism of SSDs and thus leads to higher throughput \cite{Zhuang2016}.

\item C-string manipulation functions are often significantly faster than their C++ pendants. For example, locating substrings with \texttt{strstr} is around five times faster than using the C++ \texttt{std::string} function \texttt{find}.

\item Hardcoding regular expressions with \emph{while, for, switch} or \emph{if-else} statements results in faster execution times than using standard RegEx libraries, where regular expressions are compiled at runtime into state machines.

\item Changing strings in place, instead of treating them as immutable objects, eliminates allocation and copying overhead.

\end{itemize}

\section{Experiments}

Table~\ref{tab:results} shows the running times of the resolution step of the five best placed teams.

\begin{table}[htbp]
  \caption{Comparison of the F-measure and the running times of the resolution step of the five best placed teams. The input data for the resolution step consisted of 29{,}787 in JSON formatted e-commerce websites. Measurements were taken on a
laptop running Ubuntu 19.04 with 16 GB of RAM and two Intel Core i5-4310U CPUs. The underlying SSD was a 500\,GB 860 EVO mSATA. We cleared the page cache, dentries, and inodes before each run to avoid reading the input data from RAM instead of the SSD.}
  \label{tab:results}
\resizebox{\columnwidth}{!}{
  \begin{tabular}{lcrr}
    \toprule
    Team& Language & F-measure & Running time (s)\\
    \midrule
	PictureMe (\textbf{this paper}) &C++& 0.99 & \textbf{0.61}\\
    DBGroup@UniMoRe &Python& 0.99 & 10.65\\
    DBGroup@SUSTech &C++& 0.99 & 22.13\\
    eats\_shoots\_and\_leaves &Python& 0.99 & 28.66\\
    DBTHU &Python& 0.99& 63.21\\
  \bottomrule
\end{tabular}
}
\end{table}

\section{Conclusions}

%%
%% The next two lines define the bibliography style to be used, and
%% the bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{literature}

% @inproceedings{peuwnuan2016modified,
%   title={Modified adaptive thresholding using integral image},
%   author={Peuwnuan, Kittipop and Woraratpanya, Kuntpong and Pasupa, Kitsuchart},
%   booktitle={2016 13th International Joint Conference on Computer Science and Software Engineering (JCSSE)},
%   pages={1--5},
%   year={2016},
%   organization={IEEE}
% }

\end{document}
\endinput
%%
%% End of file `sample-sigconf.tex'.

### 2022-11-30 06_Code/autovec/autovec.cpp

@@ -1,0 +1,48 @@
+#include "aligned_allocator.h"
+#include <immintrin.h>
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+// multiply each int in the array arr by num
+void multiply_by_number(int *arr, int n, int num) {
+  for (int i = 0; i < n; ++i) {
+    arr[i] *= num;
+  }
+}
+
+// no speed up in writing ugly not portable functions like this one!!!
+void multiply_by_number_avx2(int *arr, int n, int num) {
+  // int *arr should be at least 32 byte aligned
+  auto *vecs = (__m256i *)arr;
+  const auto v_mult = _mm256_set1_epi32(num);
+  // stepping vectorwise (each vector contains 8 ints)
+  for (int i = 0; i < n / 8; ++i) {
+    auto result = _mm256_mullo_epi32(vecs[i], v_mult);
+    vecs[i] = result;
+  }
+}
+
+int main() {
+  using namespace std;
+  int n = 1000000000;
+  // create two memory aligned vectors with each n ints, all ints are 3
+  aligned_vector<int> vec1(n, 3);
+  aligned_vector<int> vec2(n, 3);
+
+  auto start = omp_get_wtime();
+  multiply_by_number(vec1.data(), n, 7);
+  cout << "multiply_by_number " << omp_get_wtime() - start << " seconds" << endl;
+
+  start = omp_get_wtime();
+  multiply_by_number_avx2(vec2.data(), n, 7);
+  cout << "multiply_by_number_avx " << omp_get_wtime() - start << " seconds" << endl;
+
+  if (vec1 != vec2) {
+    cerr << "failed!\n";
+  }
+}

### 2022-11-30 06_Code/autovec/cmake-build-gcc/CMakeFiles/autovec.dir/compiler_depend.ts

@@ -1,0 +1,2 @@
+# CMAKE generated file: DO NOT EDIT!
+# Timestamp file for compiler generated dependencies management for autovec.

### 2022-11-30 06_Code/declare_simd/declare_simd.cpp

@@ -1,0 +1,58 @@
+#include "aligned_allocator.h"
+#include <algorithm>
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+// find min in array a under the condition a[i] > threshold
+int find_min_std(int *a, int n, int threshold) {
+  int result = INT32_MAX;
+  for (int i = 0; i < n; i++) {
+    if (a[i] > threshold)
+      result = std::min(a[i], result);
+  }
+  return result;
+}
+
+// "omp declare simd" tells the compiler to create several versions
+// of the function, some with vector instructions and one without
+// clang shows message that it can't create the simd versions!
+#pragma omp declare simd
+int my_min(int a, int b) { return a < b ? a : b; }
+
+// find min in array a under the condition that a[i] > threshold
+int find_min(int *a, int n, int threshold) {
+  int result = INT32_MAX;
+#pragma omp simd reduction(min : result) // min reduction
+  for (int i = 0; i < n; i++) {
+    bool is_greater = a[i] > threshold;
+    result = my_min(a[i], result) * is_greater + !is_greater * result;
+  }
+  return result;
+}
+
+int main() {
+  using namespace std;
+  int n = 500000000;
+  // create two memory aligned vectors with each n doubles, all doubles are 1.0
+  std::vector<int> a(n);
+  // quickly, create bad random numbers...
+  std::generate(a.begin(), a.end(), rand);
+  int num = rand(); // bad random number generator!! don't use it
+
+  auto start = omp_get_wtime();
+  int result1 = find_min_std(a.data(), n, num);
+  cout << "no hints, min: " << omp_get_wtime() - start << " seconds" << endl;
+
+  start = omp_get_wtime();
+  int result2 = find_min(a.data(), n, num);
+  cout << "simd hints, min: " << omp_get_wtime() - start << " seconds" << endl;
+
+  if (result1 != result2) {
+    cout << "failed!!!" << endl;
+  }
+}

### 2022-11-30 06_Code/hint/openmp_hint.cpp

@@ -1,0 +1,47 @@
+#include "aligned_allocator.h"
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+// sum up all values in arr and return sum
+double sum(const double *arr, int n) {
+  double result = 0;
+  for (int i = 0; i < n; i++) {
+    result = result + arr[i];
+  }
+  return result;
+}
+
+// sum up all values in arr and return sum, use simd hint
+double sum_with_simd_hint(const double *arr, int n) {
+  double result = 0;
+#pragma omp simd reduction(+ : result) // simd in OpenMP
+  for (int i = 0; i < n; i++) {
+    result = result + arr[i];
+  }
+  return result;
+}
+
+int main() {
+  using namespace std;
+  int n = 500000000;
+  // create two memory aligned vectors with each n doubles, all doubles are 1.0
+  aligned_vector<double> a(n, 1.0);
+  aligned_vector<double> b(n, 1.0);
+
+  auto start = omp_get_wtime();
+  auto result1 = sum(a.data(), n);
+  cout << "sum: " << omp_get_wtime() - start << " seconds" << endl;
+
+  start = omp_get_wtime();
+  auto result2 = sum_with_simd_hint(b.data(), n);
+  cout << "sum with hint: " << omp_get_wtime() - start << " seconds" << endl;
+
+  if (abs(result1 - result2) > 1e-9) {
+    cout << "failed!!!" << endl;
+  }
+}

### 2022-11-30 06_Code/pointer_aliasing/pointer_aliasing.cpp

@@ -1,0 +1,70 @@
+#include "aligned_allocator.h"
+#include <iostream>
+#include <memory>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+// sum all values in arr to *dest
+void sum_all(const int *arr, int *dest, int n) {
+  *dest = 0.0;
+  for (int i = 0; i < n; i++) {
+    *dest = *dest + arr[i]; // arr and dest may point to same value
+  }
+}
+
+// restricted pointers signal a non-overlapping memory location
+void sum_all_not_aliased(const int *__restrict__ arr,
+                               int *__restrict__ dest, int n) {
+  *dest = 0.0; // we could also create a temporary variable "tmp_dest"
+  for (int i = 0; i < n; i++) {
+    // arr and dest are not allowed to point to the same value
+    *dest = *dest + arr[i];
+  }
+}
+
+// sum all values in arr to *dest, use local variable
+void sum_all_not_aliased2(const int *arr, int *dest, int n) {
+  int tmp_dest = 0.0; // local variable for summation
+  // #pragma omp simd aligned(arr: 64) // may speed up further
+  for (int i = 0; i < n; i++) {
+    tmp_dest = tmp_dest + arr[i];
+  }
+  *dest = tmp_dest;
+}
+
+int main() {
+  using namespace std;
+  int _n = 96;
+  // allocating on the heap, prevent compiler from optimizing possible pointer
+  // aliasing away
+  auto n = std::unique_ptr<int>(new int(_n));
+  auto result = std::unique_ptr<int>(new int(0));
+  // create one memory aligned vector with n ints, all ints are 1
+  aligned_vector<int> a(*n, 1);
+
+  auto start = omp_get_wtime();
+  long final_result = 0;
+  long iterations = 100000000;
+  for (long i = 0; i < iterations; ++i) {
+    sum_all(a.data(), result.get(), *n);
+    final_result += *result;
+  }
+  cout << "possible pointer aliasing: " << omp_get_wtime() - start << " seconds"
+       << endl;
+
+  start = omp_get_wtime();
+  for (long i = 0; i < iterations; ++i) {
+    sum_all_not_aliased(a.data(), result.get(), *n);
+    final_result += *result;
+  }
[…]

### 2022-11-30 06_Code/soa_vs_aos/soa_vs_aos.cpp

@@ -1,0 +1,98 @@
+#include "aligned_allocator.h"
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+using namespace std;
+
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+// point class for constructing arrays of struct
+template <class T> struct Point {
+  T x;
+  T y;
+  T z;
+};
+
+// structure of arrays
+template <class T> class Points {
+public:
+  T *x;
+  T *y;
+  T *z;
+
+  size_t size() noexcept { return _size; }
+
+  void set_point(const T &x_coord, const T &y_coord, const T &z_coord,
+                 const size_t index) {
+    x[index] = x_coord;
+    y[index] = y_coord;
+    z[index] = z_coord;
+  }
+
+  explicit Points(size_t size) : _size(size) {
+    _x.reserve(size);
+    _y.reserve(size);
+    _z.reserve(size);
+    x = _x.data();
+    y = _y.data();
+    z = _z.data();
+  }
+
+private:
+  aligned_vector<T> _x;
+  aligned_vector<T> _y;
+  aligned_vector<T> _z;
+  size_t _size;
+};
+
+int main(int argc, char **argv) {
+  int size = 512;
+
+  auto a = aligned_vector<Point<float>>(size);
+  auto b = aligned_vector<Point<float>>(size);
+  auto squared_distances = aligned_vector<float>(size);
+
+  auto a_soa = Points<float>(size);
+  auto b_soa = Points<float>(size);
+  auto squared_distances_soa = aligned_vector<float>(size);
+
+  // initialize points
+  for (int i = 0; i < size; ++i) {
+    auto _i = float(i);
+    a[i] = Point<float>{_i + 1.0f, _i + 2.0f, _i + 3.0f};
+    b[i] = Point<float>{_i + 2.0f, _i + 3.0f, _i + 4.0f};
+    a_soa.set_point(_i + 1.0f, _i + 2.0f, _i + 3.0f, i);
+    b_soa.set_point(_i + 2.0f, _i + 3.0f, _i + 4.0f, i);
+  }
+
+  auto start = omp_get_wtime();
+  // compute the square of the distance between pairs of points (array of structs)
+  for (int j = 0; j < 10000000; ++j) {
+    for (int i = 0; i < size; ++i) {
+      float dx = a[i].x - b[i].x;
+      float dy = a[i].y - b[i].y;
[…]

### 2022-11-30 06_Code/strided_access/strided_access.cpp

@@ -1,0 +1,49 @@
+#include "aligned_allocator.h"
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+// sum up every eighth element of arr
+double sum1(const double *arr, int n) {
+  double result = 0;
+  for (int i = 0; i < n; i += 8) { // increment by 8
+    result = result + arr[i];
+  }
+  return result;
+}
+
+// sum up all elements of arr
+// using -Ofast -ffast-math compile flag --> compiler vectorizes code
+double sum2(const double *arr, int n) {
+  double result = 0;
+  for (int i = 0; i < n; ++i) { // increment by 1
+    result = result + arr[i];
+  }
+  return result;
+}
+
+int main() {
+  using namespace std;
+  int n = 500000000;
+  // create two memory aligned vectors with each n doubles, all doubles are 1.0
+  aligned_vector<double> a(n, 1.0);
+  aligned_vector<double> b(n, 1.0);
+
+  auto start = omp_get_wtime();
+  auto result1 = sum1(a.data(), n);
+  cout << "sum every eighth element: " << omp_get_wtime() - start << " seconds"
+       << endl;
+
+  start = omp_get_wtime();
+  auto result2 = sum2(b.data(), n);
+  cout << "sum every all elements: " << omp_get_wtime() - start << " seconds"
+       << endl;
+
+  if(abs(result1 - result2 / 8) > 1e-9){
+    cout << "failed!!!" << endl;
+  }
+}

### 2022-12-07 Assignments/paper.tex
add assignment 7

@@
 
+% (7)
+\section{Assignment}
+
+\subsection{Explain three vectorization clauses of your choice that can be used with \#pragma omp simd.}
+\#pragma omp simd transform a loop into a SIMD loop.
+
+Vectorization clauses with \#pragma omp simd:
+\begin{itemize}
+	\item safelen - The safelen clause allows to define a boundary for the length of vectors used 
+	within the loop. It is used to ensure the correctness of the program.
+	\item aligned - The aligned clause is used to specify the alignment (for example 64 bytes) for 
+	the pointers used within the loop.
+	notify the compiler within the loop that the pointers 
+	\item collapse - The collapse clause is used to combine multiple for loops into one.
+\end{itemize}
+
+\subsection{Give reasons that speak for and against vectorization with intrinsics compared to guided vectorization with OpenMP.}
+Vectorization with intrinsics are used to manage vectorization manually. They are basically wrappers around the corresponding assembly instructions.
+Guided vectorization with OpenMP are easier to use but no guarantee that the compiler can vectorize the codde.
+
+\subsection{What are the advantages of vector intrinsics over assembly code?}
+Vector intrinsics are easier to use than writting assembly code. One vector intrinsic can often include more than one
+assembly instruction.
+
+\subsection{What are the corresponding vectors of the three intrinsic data types: \_\_m256, \_\_m256d and \_\_m256i.}
+The intrinsic data types can hold floats, doubles or integers. Integer vectors can be interpreted as bytes (8 bits), shorts (16 bits), 
+integers (32 bits), longs (64 bits) or quads (128 bits).
+
+\_\_m256 - a vector of eight 32-bit floating point values
+\_\_m256d - a vector of four 64-bit double values
+\_\_m256i - a vector of signed or unsigned integer values
+
 %

### 2022-12-07 07_Code/aligned/aligned.cpp

@@ -1,0 +1,30 @@
+#include "aligned_allocator.h"
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+//using aligned_vector = std::vector<T>;
+
+int main() {
+  using namespace std;
+  int n = 1000;
+  int iterations = 10000000;
+
+  aligned_vector<float> a_vec(n, 1.1f);
+  aligned_vector<float> b_vec(n, 1.2f);
+  float* a = a_vec.data(); // a is 64-byte aligned
+  float* b = b_vec.data(); // b is 64-byte aligned
+  float s = 1.3f;
+
+  auto start = omp_get_wtime();
+  while(iterations --> 0) {
+#pragma omp simd aligned(a, b: 64)
+    for (int i = 0; i < n; i++) {
+      a[i] += s * b[i];
+    }
+  };
+  cout << omp_get_wtime() - start << " seconds" << endl;
+}

### 2022-12-07 07_Code/simple_for_loop/simple_for_loop.cpp

@@ -1,0 +1,81 @@
+#include "aligned_allocator.h"
+#include <immintrin.h>
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+// aligned_vector is a 64 byte aligned std::vector
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+//using aligned_vector = std::vector<T>;
+
+// pointers are not restricted
+void dummy_func_slow(int *a, int *b, const int *c, int n) {
+  for (int i = 0; i < n - 1; ++i) {
+    a[i] += b[i]; // b[i] is dependent on previous b[i + 1]
+    b[i + 1] += c[i];
+  }
+}
+
+void dummy_func_fast(int *__restrict__ a, int *__restrict__ b,
+                     const int *__restrict__ c, int n) {
+#pragma omp simd
+  for (int i = 0; i < n - 1; ++i) {
+    b[i + 1] += c[i];
+  }
+#pragma omp simd
+  for (int i = 0; i < n - 1; ++i) {
+    a[i] += b[i];
+  }
+}
+
+void dummy_func_fast2(int *__restrict__ a, int *__restrict__ b,
+                      const int *__restrict__ c, int n) {
+  a[0] += b[0];
+#pragma omp simd aligned(a, b, c: 64)
+  for (int i = 1; i < n - 1; ++i) {
+    b[i] += c[i - 1];
+    a[i] += b[i];
+  }
+  b[n - 1] += c[n - 2];
+}
+
+int main() {
+  using namespace std;
+  int n = 1000;
+  int iterations = 1000000;
+
+  aligned_vector<int> a(n, 1);
+  aligned_vector<int> b(n, 2);
+  aligned_vector<int> c(n, 3);
+  auto start = omp_get_wtime();
+  // don't care here with integer overflow, just dummy computation
+  for (int i = 0; i < iterations; ++i) {
+    dummy_func_slow(a.data(), b.data(), c.data(), n);
+  };
+  cout << "dummy slow: " << omp_get_wtime() - start << " seconds" << endl;
+
+  aligned_vector<int> a2(n, 1);
+  aligned_vector<int> b2(n, 2);
+  aligned_vector<int> c2(n, 3);
+  start = omp_get_wtime();
+  // don't care here with integer overflow, just dummy computation
+  for (int i = 0; i < iterations; ++i) {
[…]

### 2022-12-07 07_Code/thread_simd/thread_simd.cpp

@@ -1,0 +1,24 @@
+#include <iomanip>
+#include <iostream>
+#include <omp.h>
+
+using namespace std;
+
+int main() {
+  int num_steps = 100000000; // amount of rectangles
+  double width = 1.0 / double(num_steps); // width of a rectangle
+  double sum = 0.0; // for summing up all heights of rectangles
+
+  double start_time = omp_get_wtime(); // wall clock time in seconds
+
+#pragma omp parallel for simd reduction(+:sum) schedule(simd:static)
+  for (int i = 0; i < num_steps; i++) {
+    double x = (i + 0.5) * width; // midpoint
+    sum = sum + (1.0 / (1.0 + x * x)); // add new height of a rectangle
+  }
+  double pi = sum * 4 * width; // compute pi
+  double run_time = omp_get_wtime() - start_time;
+
+  cout << "pi with " << num_steps << " steps is " << setprecision(17)
+       << pi << " in " << setprecision(6) << run_time << " seconds\n";
+}

### 2022-12-07 08_Code/demo_fma.cpp

@@ -1,0 +1,17 @@
+#include <immintrin.h>
+#include <iostream>
+
+int main() {
+  // initialize float vectors
+  __m256 a = _mm256_setr_ps(1, 2, 3, 4, 5, 6, 7, 8);
+  __m256 b = _mm256_setr_ps(2, 3, 4, 5, 6, 7, 8, 9);
+  __m256 c = _mm256_setr_ps(1, 2, 1, 2, 1, 2, 1, 2);
+
+  // compute fused-multiply-add with vectors
+  auto dst = _mm256_fmadd_ps(a, b, c);
+
+  // show results --> 3 8 13 22 31 44 57 74 
+  auto arr = (float *)&dst;
+  for (int i = 0; i < 8; ++i)
+    std::cout << arr[i] << " ";
+}

### 2022-12-07 08_Code/demo_permutation.cpp

@@ -1,0 +1,21 @@
+#include <immintrin.h>
+#include <iostream>
+
+int main() {
+  float arr[8] = {1, 2, 3, 4, 5, 6, 7, 8};
+
+  // load (unaligned) array data into vector register
+  __m256 v = _mm256_loadu_ps(arr);
+
+  // indices of permutation
+  __m256i idx = _mm256_setr_epi32(7, 6, 5, 4, 3, 2, 1, 0);
+  // reverse vector (we can permute the vector in any order)
+  __m256 result = _mm256_permutevar8x32_ps(v, idx);
+  // _mm256_permutevar8x32_ps (latency: 3, throughput: 1)
+
+  // store data from vector register into array (unaligned)
+  _mm256_storeu_ps(arr, result);
+
+  // show results: 8, 7, 6, 5, 4, 3, 2, 1
+  for (float i : arr) std::cout << i << " ";
+}

### 2022-12-07 08_Code/demo_shuffle.cpp

@@ -1,0 +1,20 @@
+#include <immintrin.h>
+#include <iostream>
+
+int main() {
+  // initialize float vector
+  __m256 v = _mm256_setr_ps(1, 2, 3, 4, 5, 6, 7, 8);
+
+  // shuffle float vector v into a new vector result
+  // 1, 2, 3, 4, 5, 6, 7, 8 -->  2, 1, 4, 3, 6, 5, 8, 7
+  __m256 result = _mm256_shuffle_ps(v, v, 0b10110001); 
+  // _mm256_shuffle_ps (latency: 1, throughput: 1)
+  // is one of the most difficult intrinsic functions to understand
+  // we could also do the same shuffle with:
+  // _mm256_permutevar_ps(v, _mm256_setr_epi32(1, 0, 3, 2, 1, 0, 3, 2));
+
+  // show results
+  auto arr = (float *)&result;
+  for (int i = 0; i < 8; ++i)
+    std::cout << arr[i] << " ";
+}

### 2022-12-07 08_Code/partition_vector.cpp

@@ -1,0 +1,292 @@
+#include <array>
+#include <immintrin.h>
+#include <iostream>
+
+const __m256i permutation_masks[256] = {
+    _mm256_setr_epi32(0, 1, 2, 3, 4, 5, 6, 7),
+    _mm256_setr_epi32(1, 2, 3, 4, 5, 6, 7, 0),
+    _mm256_setr_epi32(0, 2, 3, 4, 5, 6, 7, 1),
+    _mm256_setr_epi32(2, 3, 4, 5, 6, 7, 0, 1),
+    _mm256_setr_epi32(0, 1, 3, 4, 5, 6, 7, 2),
+    _mm256_setr_epi32(1, 3, 4, 5, 6, 7, 0, 2),
+    _mm256_setr_epi32(0, 3, 4, 5, 6, 7, 1, 2),
+    _mm256_setr_epi32(3, 4, 5, 6, 7, 0, 1, 2),
+    _mm256_setr_epi32(0, 1, 2, 4, 5, 6, 7, 3),
+    _mm256_setr_epi32(1, 2, 4, 5, 6, 7, 0, 3),
+    _mm256_setr_epi32(0, 2, 4, 5, 6, 7, 1, 3),
+    _mm256_setr_epi32(2, 4, 5, 6, 7, 0, 1, 3),
+    _mm256_setr_epi32(0, 1, 4, 5, 6, 7, 2, 3),
+    _mm256_setr_epi32(1, 4, 5, 6, 7, 0, 2, 3),
+    _mm256_setr_epi32(0, 4, 5, 6, 7, 1, 2, 3),
+    _mm256_setr_epi32(4, 5, 6, 7, 0, 1, 2, 3),
+    _mm256_setr_epi32(0, 1, 2, 3, 5, 6, 7, 4),
+    _mm256_setr_epi32(1, 2, 3, 5, 6, 7, 0, 4),
+    _mm256_setr_epi32(0, 2, 3, 5, 6, 7, 1, 4),
+    _mm256_setr_epi32(2, 3, 5, 6, 7, 0, 1, 4),
+    _mm256_setr_epi32(0, 1, 3, 5, 6, 7, 2, 4),
+    _mm256_setr_epi32(1, 3, 5, 6, 7, 0, 2, 4),
+    _mm256_setr_epi32(0, 3, 5, 6, 7, 1, 2, 4),
[…]

### 2022-12-07 08_Code/random_numbers.cpp

@@ -1,0 +1,49 @@
+#include <immintrin.h>
+#include <iostream>
+
+/********** scalar Xoroshiro128+ **********/
+inline uint64_t rotl(const uint64_t x, int k) {
+  return (x << k) | (x >> (64 - k)); // rotate
+}
+
+inline uint64_t next(uint64_t &s0, uint64_t &s1) {
+  s1 ^= s0;                            // modify s1
+  s0 = rotl(s0, 24) ^ s1 ^ (s1 << 16); // modify s0
+  s1 = rotl(s1, 37);                   // modify s1
+  return s0 + s1;
+}
+
+/********** vectorized Xoroshiro128+ **********/
+#define VROTL(x, k)                                          \
+  _mm256_or_si256(_mm256_slli_epi64((x), (k)),               \
+                  _mm256_srli_epi64((x), 64 - (k))) // rotate
+
+inline __m256i vnext(__m256i &s0, __m256i &s1) {
+  s1 = _mm256_xor_si256(s0, s1);                    // modify s1
+  s0 = _mm256_xor_si256(_mm256_xor_si256(VROTL(s0, 24), s1),
+                        _mm256_slli_epi64(s1, 16)); // modify s0
+  s1 = VROTL(s1, 37);                               // modify s1
+  return _mm256_add_epi64(s0, s1);
+}
+
+int main() {
+  {
+    uint64_t s0 = 1314472907419283471; // seed 0
+    uint64_t s1 = 7870872464127966567; // seed 1
+    for (int i = 0; i < 4; ++i) {
+      std::cout << next(s0, s1) << std::endl;
+    }
+  }
+  std::cout << std::endl;
+  {
+    auto s0 = _mm256_setr_epi64x(8265987198341093849, 3762817312854612374,
+                                 1324281658759788278, 6214952190349879213); // vector seed 0
+    auto s1 = _mm256_setr_epi64x(2874178529384792648, 1257248936691237653,
+                                 7874578921548791257, 1998265912745817298); // vector seed 1
+    auto random_vec = vnext(s0, s1);
+    for (int i = 0; i < 4; ++i) {
+      std::cout << ((uint64_t *)&random_vec)[i] << std::endl;
+    }
+    // generate next random vector with vnext(s0, s1)
+  }
+}

### 2023-02-15 Assignments/paper.tex
update paper

@@
 
-\_\_m256 - a vector of eight 32-bit floating point values
-\_\_m256d - a vector of four 64-bit double values
-\_\_m256i - a vector of signed or unsigned integer values
+\begin{itemize}
+	\item \_\_m256 - a vector of eight 32-bit floating point values 
+	\item \_\_m256d - a vector of four 64-bit double values
+	\item \_\_m256i - a vector of signed or unsigned integer values
+\end{itemize}
+
+% (8)
+\section{Assignment}
+% Vector intrinsics, performance portable code, manually vectorize algorithms
+
+\subsection{Explain the naming conventions for intrinsic functions.}
+Naming conventions: \_<vector\_size>\_<operations>\_<suffix> \\
+\_<vector\_size> - specifies the size of the vector returned by the intrinsic function \\
+\_<operations> - specifies the operation, like add, sub, mult \\
+\_<suffix> - specifies the data type of the input arguments \\
+Example: \_\_m256d \_mm256\_add\_pd (\_\_m256d a, \_\_m256d b) \\
+Vector size: m256d - 64-bit floating point values \\
+Operation: add - add the corresponding elements of two vectors \\
+Suffix: pd - double precision (epi, epu - signed or unsigned)
+
+\subsection{What do the metrics latency and throughput tell you about the performance of an intrinsic function?}
+% metrics latency and throughput about performance
+Latency is the number of cycles an intrinsic takes until its result is available. 
+Throughput specifies how many cylces it takes to start the next intrinsic function of the same kind.
+Latency and throughput can vary depending on the underlying CPU. For the add intrinsic latency varies from 2 to 4
+and the throughput is 0.5. A latency of 4 means that it takes 4 cycles until the result is available. A throughput of 0.5
+means that within one cycle 2 instructions from the intrinsic can be started. In general there are two units on the CPU to
+compute such functions. Because there are 2 units the throughput is 0.5.
+Within one cycle two add intrinsic functions can be started. It takes up to 4 cycles until a result is available.
+% Software pipelining
+
+\subsection{How do modern processors realize instruction-level parallelism?}
+Instruction-level parallelism (superscalar) refers to several instruction which are being evaluated simultaneously on
+different functional units.
+Multiple instructions can be executed at the same time within a cycle using these different functional units but one CPU core.
+
+\subsection{How may loop unrolling affect the execution time of compiled code?}
+% loop unrolling
+Loop unrolling is a technique to make use of instruction-level parallelism.
+It can be used for scalar or vector operations.
+The body of the loop is replicated and the logic is adjusted. The loop is unrolled by a factor k, 
+which determines the number of times the body is replicated.
+Loop unrolling may lead to an incresed demand for registers depending on the factor used. 
+It can increase compile time, the program size because the body is replicated and loop unrolling needs additional space in the instruction cache.
+Loop unrolling is more affective when using vector instructions to exploit instruction-level parallelism.
+
+\subsection{What does a high IPC value (instructions per cycle) mean in terms of the performance of an algorithm?}
+% high IPC value (instructions per cycle) for performance
+IPC (instructions per cycle) is an aspect of the processors's performance referring to the average number of instructions
+executed in one clock cylce. It is one indication for a systems performance. 
+A high IPC indicates that the CPU is efficiently used.
+To calculate the instructions per cycle the number of instructions is divided by the cycles.
+
+% (9)
+\section{Assignment}
+
+\subsection{How do bandwidth-bound computations differ from compute-bound computations?}
+Bandwith-bound computations have memory access as bottleneck. When loading data main memory is accessed.
+For compute-bound computations the bottleneck is the CPU and main memory or RAM are not accessed.
+When optimizing the cache usage, bandwith-bound computations show more effective results.
+
+% bandwidth-bound computations - bottleneck memory access
+% Main memory is accessed
+% Elements stored in RAM are accessed
+% When optimizing cache usage, bandwith-bound computations are more effective.
+
+% compute-bound computations - bottleneck CPU
+% Main memory, RAM is not accessed
+% Vector instructoins are more relevant
+
+% SRAM - Cache, latency is lower than DRAM, closer to CPU
+% DRAM - Arbeitsspeicher, loads program data, has to be refreshed periodically
+
+% write programs that make use of the cache effectively and make less use of DRAM
+
+% Cache line - Intel 64 bytes - 8 double precision values, 16 single precision values, 64 bytes
+% M1 - 128 bytes - 16 double precision values, 32 single precision values, 128 bytes
+
+\subsection{Explain why temporal locality and spatial locality can improve program performance.}
+In modern computers caches are used to reduce the processor-memory gap.
+For this to work programs need to establish good temporal and spatial locality.
+The cache holds elements for the CPU that are often accessed to avoid access to main memory.
+The CPU can access the data directly through the cache.
+Spatial locality means that elements should be stored efficiently. 
+If one element of a cache line is accessed the whole cache line is loaded.
+Therefore, programs should be written in such a way that the other elements are used aswell.
+Everything that is loaded should be used as efficiently as possible.
+Temporal locality means that programs should be written in such a way that the data is used as often as possible.
+Blocking is a mechanism which makes use of temporal locality. 
+The computation is divided into blocks that are united afterwards.
+
+\subsection{What are the differences between data-oriented design and object-oriented design?}
+Data-oriented design (DOD) is a way to design programs that make efficient use of CPU cache. 
+DOD modifies the input to obtain the desired output.
+In object-oriented design everything revolves around the defining, producing and operating on objects.
+Objects interact with each other through functions.
+In data-oriented design functionality and data is seperated. Functions are general purpose,
+It shifts the perspective onto the data and how it is laid out in memory and read from there.
+DOD uses Structures of Arrays (SoA) instead of Array of Structures (AoS) for data that is often used.
+
+\subsection{What are streaming stores?}
+Streaming stores are a way to write data from the CPU directly to random access memory (RAM) and not not through cache.
+Data is not written to cache.
+It can lead to better performance, if the data is not used.
+Another way to use streams is streaming loads to load data directly from RAM to the CPU.
+
+\subsection{Describe a typical cache hierarchy used in Intel CPUs.}
+Modern Intel CPUs typically have an L1 and L2 cache within each core. 
+The L1 cache is divided into an L1 data cache and an L1 instruction cache.
+L2 is the unified cache for data and instructions. \\
+Outside of the CPU cores there is a L3 cache, which is shared by all cores and represents the interface to main memory.
+The caches size goes from the smallest L1 to the largest L3 cache.
+\\
+Each cache can be thought of as a hash table with for example 64 rows. Each row has a number of buckets,
+where each bucket is a cache line with a size of 64 bytes for Intel CPUs.
+
+% Read throughpput: how many MB of data can be moved to the CPU within one second
+
+\subsection{What are cache conflicts?}
+Caches are basically hardware hash tables and as such they suffer from conflicts.
+Each hash bucket can contain N cache lines.
+Cache conflicts occur when more cache lines are put into a bucket than it can hold.
+
+% (10)
+\section{Assignment}
+
+\subsection{Name and explain some useful compiler flags during development.}
+\begin{itemize}
+	\item Wall - enables compiler's warning messages
+	\item g - generates debug information
+	\item fsanitize=address - decets out-of-bounds-access, use-after-free and memory leaks
+	\item fsanitize=undefined - detects undefined behavior at runtime
+\end{itemize}
+
+\subsection{How could Intel oneAPI help you write better programs?}
+Intel oneAPI is a toolkit containing Intel compilers supporting CPUs,GPUs, FPGAs and other accelerators.
+It provides a cross-architecture programming model including performance libraries, analyzer and debugger tools as well as 
+domain-specific toolkits.
+
+\subsection{What can we learn from the following quote? Premature optimization is the root of all evil (Donald Knuth).}
+Premature optimization refers to blindly optimizing everything without thinking about it.
+It is better to first profile and find bottlenecks that can be optimized.
+The approach shouldn't be to invasive especially in a program written by others.
+It is better to keep things simple.
+
+% (11)
+\section{Assignment}
+
+\subsection{What is Cython?}
+Cython is a superset of Python. 
+It is used for computations that could slow down performance.
+Python code tends to be slow. 
+Cython is used for critical computations to speed up perfromance.
+It is an interface with C and C++ libraries.
+
+\subsection{Describe an approach how Python programs can be accelerated with the help of Cython.}
+
+\subsection{Describe two ways for compiling a .pyx Cython module.}
+To compile a .pyx Cython module the command cythonize can be used. 
+It compiles the file into a C/C++ file.
+After that the file is compiled into an extension module that is importable from Python.
+\\
+Another way to compile a .pyx file is to use pyximport. 
+This is used to automatically recompile and reload the module.
+The line "import pyximport" and "pyximport.install()" is used at the start of the Python script.
+
+\subsection{Name and describe two compiler directives in Cython.}
+% cython: language_level = 3
+% cython: boundscheck = False
+% cython: wraparound = False
+% cytho: cdivision = True
+
+\subsection{What is the difference between, def, cdef, and cpdef when declaring a Cython function?}
+\begin{itemize}
+	\item def - normal python function
+	\item cdef - can be called only in cython
+	\item cpdef - can be called in cython and python
+\end{itemize}
+
+\subsection{What are typed memoryviews especially useful for in Cython?}
+Memoryviews are useful for passing Python of Numpy arrays to Cython.
+
+% (12)
+\section{Assignment}
+
+\subsection{What are extension types in the context of Python?}
+Extension types in can be used to wrap C / C++ code looking like Python objects.
+They can be used to boost performance. Cython makes it easy to create extension types.
+
+\subsection{How do extension types data fields in Cython differ from data fields in Python classes?}
+Extension type data fileds in Cython can be accessed only within Cython code.
+To access fields from Python an access level - readonly of public - for that field has to be provided.
+
+\subsection{Give a simple description of how to wrap C / C++ code in cython?}
+We have a C / C++ library we want to use in Pythong. 
+In our Python code we include the library using \#include "pi.h".
+In Cython we describe all we need with compiler directives.
+
+% (13)
+\section{Assignment}
+
+\subsection{Delimit from each other the following SSD parts: Cells, Pages and Blocks.}
+In SSDs cells are used to store digital bits. 
+Cells are NAND based and can store from one to four bits depending on the type of NAND.
+A group of cells forms a page. The smallest unit to read or write with SSDs is a page.
+It can only be written once. This means that there are no operations to overwrite a page.
+When data is updated it is written to another page. 
+The old page is marked as invalid until it is erased.
+Blocks are usually 512KB, 1MB (about 128 or 256 pages). The smallest unit to erase is a block.
+
+\subsection{What is the purpose of garbage collection in SSDs?}
+A block contains pages. The number of invalid pages is compared to a threshold. 
+If it exceeds the threshold, the block is claimed by garbage collection.
+Furthermore, garbage collection moves related data segments next to each other.
+
+\subsection{What is the purpose of wear leveling in SSDs?}
+War leveling is used to prolong the life of an SSD.
+
+\subsection{Tell some interesting things about SSDs with an M.2 form factor.}
+The Host Interface provides an interface between host and SSD, such as PCI, SATA or SAS.
+To efficiently communicate with the interface the protocol AHCI or NVMe is used.
+NVMe protocols are more efficiently with SATA interface.
+The M.2 form factor exists for PCIe and SATA interfaces.
+It has multiple key slots, which have to be checked for compatibility.
+Performance is higher for write operations than read operations.
+
+\subsection{What influence do garbage collection and wear leveling have on write amplification of an SSD?}
+Write amplification happens when writing data and the data to write is multiplied into a disproportional amount of data on the SSD.
+Garbage collection and wear levelling can increase write amplification.
+
+\subsection{Discuss three different recommendations for writing code for SSDs.}
+When writing code for SSDs it is important to avoid scattered updates and use compact data structures.
+Additionally, for large IO, only a few threads suffice to achieve maximum performance.
+
+\subsection{How could the CPU load for IO be reduced?}
+To reduce CPU load one can use asynchronous IO system calls instead of blocking ones.
+Furthermore, the OS buffering has to be diabled.
+
+\subsection{How could you solve problems that do not fit in DRAM without major code adjustments?}
+To solve such problems one can use memory mapped files on flash.
 

### 2023-02-19 Assignments/paper.tex
update

@@
 \subsection{Describe how parallelism differs from concurrency}
+Concurrency refers to two or more tasks being executed, but not simultaenously. 
+Parallelism is about the exection of multiple tasks at the same time.
 Parallelism is a subset of concurrency. If a system is parallel it is also concurrent, but not the other way around.
-A system is considered parallel if two or more tasks are executed simultaneously.
-\\
-Concurrency only means support for two or more actions at the same time. 
-Parallelism refers to execution of more than one action at the same time.
 
 \subsection{What is fork-join parallelism?}
-Fork-join parallelism refers to a concept where we have a master thread that divides into a team of threads. 
-With this concept we only have parallel regions where threads are executed simultaneously. 
+Fork-join parallelism is a concept where we have a master thread that divides into more threads. 
+This opens a parallel regions allowing simultaneous excecution.
 In between these parallel regions the execution is sequential following the master thread. 
@@
 With fork-join parallelism this line is our master thread. 
-Following our master thread at some point the thread can divide into two or more threads. 
+Following our master thread at some point the thread divides into two or more threads. 
 These threads are joined after some time to our master thread. 
-In these sections where we have two or more threads execution will be parallel and after these sections sequential.
+In these sections where we have two or more threads, execution will be parallel and after these sections sequential.
 
@@
 \paragraph{Discuss one thing - Caches}
-As discussed in Chapter 1 of the book caches are helpful to deal with the processor-memory gap.
-
+As explained in Chapter 1 of the book "Computer Systems: A Programmer’s Perspective" caches are helpful to deal with the processor-memory gap. 
+The processor-memory gap refers to the difference in performance between processor and memory.
 The processor can read data from the register file, which is within the CPU almost 100 times faster than from memory. 
 The gap between processor and memory is continuously growing. 
-
+\\
 Cache memory is used to bridge the gap. It is used to temporarily store information that will likely be used in the future. 
-There exist several levels of caches -  L1, L2, L3 - which go from smallest to largest.
-
+There are several levels of cache -  L1, L2, L3 - which go from smallest to largest. 
 A cache reduces the acess time to data in memory. Frequently used data and instructions are kept in the cache. 
-Access to data kept there is faster.
-
+Accessing data from the cache is faster.
+\\
 Modern desktops, servers and industrial CPUs have at least three independent caches: instruction cache, data cache and the TLB. 
@@
 \begin{itemize}
-	\item L1 cache - Primary cache is fast but small and usually embedded in the processor chip (CPU)
-	\item L2 cache - Secondary cache is larger and can be embedded in the CPU or on a seperate chip or coprocessor containing 
-	a high-speed alternative system bus connectin cache and CPU.
-	\item L3 cache - specialized memory to improve L1 and L2. L3 is usually double the speed of DRAM
+	\item L1 cache or primary cache is fast but small and usually embedded in the processor chip (CPU).
+	\item L2 cache or secondary cache is larger and can be embedded in the CPU or on a seperate chip or coprocessor containing 
+	a high-speed alternative system bus connecting cache and CPU.
+	\item L3 cache is used to improve L1 and L2. L3 is usually double the speed of DRAM.
 \end{itemize}
 The small caches L1 are backed up by larger and slower caches L2, L3 to address the tradeoff between cache latency and hit rate. 
-The faster cache is generally checked firt. If that cache misses, the next chache is checked and so on, before accessing main memory.
-
+The faster cache is generally checked first. If that cache misses, the next chache is checked and so on before accessing main memory.
+\\
 For example the ARM-base Apple M1 CPU has 8 cores: 4 high-performance and four high-efficiency cores. 
 The four high-performance cores have a 192 KiB L1 cache for each of the cores and the four high-efficiency cores only have 128 KiB.
-
+\\
 The Translation lookaside buffer is used to speed up the translation from virtual to physical addresses. 
@@
 The figure illustrates the “Top” and “Bottom” referring to computer performance gains. 
-The “Bottom” refers to the miniaturization of computer components seen in the last decades. \\
+The “Bottom” is about the miniaturization of computer components seen in the last decades. \\
 Due to physical limits the opportunities for gains at the bottom will slowly come to an end.
@@
 \subsection{Explain the differences betweent static and dynamic schedules in OpenMP.}
-Static and dynamic schedules differ in how the work \( iterations of the for loop \)) is spread across the threads. 
-Static means that it is decided at the beginning and dynamic means that it is decided at runtime. 
+Static and dynamic schedules differ in how the work (iterations of the for loop) is spread between threads. 
+\\
+Static schedules decide at the beginning and dynamic means that it is decided at runtime. 
 Each thread will work on a chunk of values and then take the next chunk that hasn't been worked on by any thread.
@@
 in case the workload varies between different iterations of the for loop.
-The chunk size can be specifies; for static schedules it is one per default.
+The chunk size can be specified; for static schedules it is one per default.
 
@@
 strd::atomic::compare\_exchange\_weak  succeeds only if the value of the variable to be updated equals the first provided argument.
-For example if for final\_solution.compare\_exchange\_weak(previous, i) final\_solution equals to previous. 
+For example if for final\_solution.compare\_exchange\_weak(previous, i) final\_solution is equal to previous. 
 Previous is updated with i, if it fails.
@@
 \subsection{How does ordered clause in OpenMP work in conjunction with a parallel loop?}
-An ordered clause is used within a parallel region. It opens a region where the execution is sequential, 
-in order to prevent race conditions.
+An ordered clause is used within a parallel region. It opens a region where the execution is sequential. 
+It is used to prevent race conditions.
 
@@
 \subsection{Explain how reductions work internally in OpenMP}
-A reduction is an operation form OpenMP that is used in parallel computing. 
-It reduces the calues into a single result.
-The syntax of the operation is: \lstinline{reduction(op:list)}. op refers to the operation, which can be +, -, *, and so on. 
+A reduction is an operation from OpenMP that is used in parallel computing. 
+It reduces the values into a single result.
+The syntax of the operation is: \lstinline{reduction(op:list)}. 
+op refers to the operation (addition, subtraction, and so on). 
 list refers to a list of variables that is reduced.
@@
 \subsection{What is the purpose of a barrier in parallel computing?}
-A barrier can be set within a parallel region. Setting a barrier means that the threads will execute the code until that specific barrier. 
-There they wait until all threads have reached the barrier and only then continue with execution.
+A barrier can be set within a parallel region. 
+Setting a barrier means that the threads will execute the code up to that specific barrier, which is a specific point in execution. 
+At this point the threads wait until all have reached the barrier and only then continue with execution.
 Barriers are used for synchronization in parallel regions, where they enforce a specific execution order. \\
@@
 \subsection{Explain the difference between the library routines: omp\_get\_num\_threads(), omp\_get\_num\_procs() and omp\_get\_max\_threads().}
-\begin{itemize}
-	\item omp\_get\_num\_threads() - 
-	The routine is called inside a parallel region. It is used to get the number of threads used within a parallel region.
-	\item omp\_get\_num\_procs() - 
-	The routine is used to get the number of logical cored. It can be used to define the number of threads reasonable to use 
-	in a parallel region. Per default the number of threads is equal to the number of logical cores.
-	\item omp\_get\_max\_threads() - 
+omp\_get\_num\_threads():
+The routine is called inside a parallel region. 
+It is used to get the number of threads used within a parallel region.
+\\
+omp\_get\_num\_procs():
+The routine is used to get the number of logical cores. 
+It can be used to define the number of threads reasonable to use in a parallel region. 
+Per default the number of threads is equal to the number of logical cores.
+\\
+omp\_get\_max\_threads():
 The routine is called outside a parallel region and returns the number of threads supported in a parallel region. 
 When the value is changed with omp\_set\_num\_threads() within a parallel region, the method returns the set value.
-\end{itemize}
 
@@
 There are three different storage attributes: shared, firstprivate and private.
-
+\\
 Private attributes make sure that each thread gets an uninitialized copy of the variable.
-Firstprivate attributes make sure that eacht thrad gets an identical initialized copy of the variable. 
+Firstprivate attributes make sure that each thread gets an identical initialized copy of the variable. 
 The variable is a local variable for each thread.
@@
 Such algorithms can be paralleized in OpenMP using the constructs taks and taskwait for the recursive call.
-OpenMP tasks are independent work blocks. A queueing system dynamically handles the assignment of threads to the tasks. 
-Each threads picks up a task from the queue and executes it. This goes on until the queue is empty. 
+OpenMP tasks are independent work blocks. 
+A queueing system dynamically handles the assignment of threads to the tasks. 
+Each thread picks up a task from the queue and executes it. 
+This goes on until the queue is empty. 
 Tasks are defined within a parallel region and within a single construct, to make sure that one thread creates the tasks.
@@
 \begin{itemize}
-	\item Tasks - With task and taskwait
+	\item Tasks - with task and taskwait
 	\item Avoid copying - to prevent that additional memory is created recursively and make sure that initially memory 
-	is instantiated and reused.
+	is instantiated and reused
 	\item if Clause - task will only be created, if a certain condition is true
@@
 To prevent reordering we can use atomic types in C or C++.
-
+\\
 Atomicity means that something can’t be divided into smaller pieces. Threads need to use atomic reads and writes to share data. 
 If they are not atomic they are so-called torn reads and writes.
-To ensure atomicity one need to make sure that variables used for thread synchronization aren’t larger than the CPU word size.
+To ensure atomicity one needs to make sure that variables used for thread synchronization aren’t larger than the CPU word size.
 
@@
 \subsection{What role do targets play in CMake?}
-Targets define what is build. They are executables and libraries.
+Targets define what is built. They are executables and libraries.
 Targets have constructors to build executables, libraries or tests.
-Targets can be thought of as objects with different properties. To change these properties we can use member functions.
-Within member functions the properties can be specified. Properties can flags, directories or linked libraries.
+Targets can be thought of as objects with different properties. 
+To change these properties we can use member functions.
+Within member functions the properties can be specified. 
+Properties can be flags, directories or linked libraries.
 
 \subsection{How would you proceed to optimize code?}
-The most important thing is to try solving the problem. Once that is done and my code works, I would look at the performance.
+The most important thing is to try solving the problem. Once that is done and the code works, I would look at the performance.
 If performance can be improved, I would look at ways to do so. I would try parallelizing sections of code, try finding ways to improve the algorithm.
@@
 \subsection{Name some characteristics of the instruction sets: SSE, AVX(2), AVX-512}
-SSE, AVX2, AVX-512 refer to instruction sets on Intel CPUs. They differ in the supported vector length and the number of registers.
-
+SSE, AVX2, AVX-512 are instruction sets on Intel CPUs. They differ in the supported vector length and the number of registers.
+\\
 SSE: 128-bit vector length, 1999 - 2009 years of launch, 16 registers
+\\
 AVX2: 256-bit vector length, 2011 / 2013 years of launch, 16 registers
+\\
 AVX-512: 512-bit vector length, 2017 year of launch, 32 registers
-
-CPUs for high-performance computing support AVX-512 instruction set, wehere as modern mainstream CPUs support AVX2 at most.
+\\
+CPUs for high-performance computing support the AVX-512 instruction set, wehereas modern mainstream CPUs support AVX2 at most.
 
@@
 To ensure safe optimizations, the compiler always assumes pointer aliasing. 
-Therefore, we can give the compiler hints, to ensure it that no memory aliasing was made.
-This can be done using the restrict keyword (\*\_\_restrict\_\_ for gcc). Memory aliasing is important for pointers as well as references.
+Therefore, we can give the compiler hints, to ensure that no memory aliasing was made.
+This can be done using the restrict keyword (\*\_\_restrict\_\_ for gcc). 
+Memory aliasing is important for pointers as well as references.
 
 \subsection{What are the advantages of unit stride (stride-1) memory access compared to accessing memory with larger strides (for example stride-8)?}
-stride-1 - access sequential elements in memory, where each fields are equally distant. The distance in this case is called stride.
+stride-1 - access sequential elements in memory, where all of the fields are equally distant to one another. The distance in this case is called stride.
 The difference between stride-1 and larger strides stride-n is that every nth element is accessed.
-
+\\
 Stride-1 is faster than larger strides. There is a maximum bandwith that can be loaded from memory. 
 Using larger strides leads to loading a higher bandwidth without fully using it.
-
+\\
 We can check this by looking at the summation of two arrays. 
@@
 Array of Structures: one array with the elements for each entry.
-
-For example when we want to save x,y,z coordinates. When using a Structure of Arrays we have one struct and three arrays for each coordinate.
-If we use an Array of Structures we have a struct array with three entries. Each entry stores the values of the x, y, z coordinates.
-
+\\
+For example when we want to save the x, y and z coordinates. When using a Structure of Arrays we have one struct and three arrays for each coordinate.
+If we use an Array of Structures we have a struct array with three entries. 
+Each entry storing the values of the x, y and z coordinates.
+\\
 A Structure of Arrays is good for vectorization. In our example with the coordinates, we can access all the entries of a specific coordinate
@@
 With a SoA the arrays are kept seperate for each structure field. Memory access is contiguous if we perform vectorization of the structure instances.
-When performing operations on one array of the structure, SoA leads better bandwidth usage.
+When performing operations on one array of the structure a SoA leads to better bandwidth usage.
 
@@
 \subsection{Explain three vectorization clauses of your choice that can be used with \#pragma omp simd.}
-\#pragma omp simd transform a loop into a SIMD loop.
-
-Vectorization clauses with \#pragma omp simd:
+\#pragma omp simd transforms a loop into a SIMD loop.
+\\
+Some vectorization clauses with \#pragma omp simd are:
 \begin{itemize}
@@
 	the pointers used within the loop.
-	notify the compiler within the loop that the pointers 
 	\item collapse - The collapse clause is used to combine multiple for loops into one.
@@
 \subsection{Give reasons that speak for and against vectorization with intrinsics compared to guided vectorization with OpenMP.}
-Vectorization with intrinsics are used to manage vectorization manually. They are basically wrappers around the corresponding assembly instructions.
-Guided vectorization with OpenMP are easier to use but no guarantee that the compiler can vectorize the codde.
+Vectorization with intrinsics is used to manage vectorization manually. 
+It is basically wrappers around the corresponding assembly instructions.
+\\
+Guided vectorization with OpenMP is easier to use but there is no guarantee that the compiler can vectorize the code.
 
 \subsection{What are the advantages of vector intrinsics over assembly code?}
-Vector intrinsics are easier to use than writting assembly code. One vector intrinsic can often include more than one
+Vector intrinsics are easier to use than writing assembly code. One vector intrinsic often consists of more than one
 assembly instruction.
@@
 \subsection{What are the corresponding vectors of the three intrinsic data types: \_\_m256, \_\_m256d and \_\_m256i.}
-The intrinsic data types can hold floats, doubles or integers. Integer vectors can be interpreted as bytes (8 bits), shorts (16 bits), 
+The intrinsic data types can hold floats, doubles or integers. Vectors can be interpreted as bytes (8 bits), shorts (16 bits), 
 integers (32 bits), longs (64 bits) or quads (128 bits).
@@
 \subsection{Explain the naming conventions for intrinsic functions.}
-Naming conventions: \_<vector\_size>\_<operations>\_<suffix> \\
-\_<vector\_size> - specifies the size of the vector returned by the intrinsic function \\
-\_<operations> - specifies the operation, like add, sub, mult \\
-\_<suffix> - specifies the data type of the input arguments \\
+
+Naming conventions: \_<vector\_size>\_<operations>\_<suffix>
+
+\begin{itemize}
+	\item \_<vector\_size>: specifies the size of the vector returned by the intrinsic function
+	\item \_<operations>: specifies the operation, like add, sub, mult
+	\item \_<suffix>: specifies the data type of the input arguments
+\end{itemize}
 Example: \_\_m256d \_mm256\_add\_pd (\_\_m256d a, \_\_m256d b) \\
@@
 Throughput specifies how many cylces it takes to start the next intrinsic function of the same kind.
-Latency and throughput can vary depending on the underlying CPU. For the add intrinsic latency varies from 2 to 4
-and the throughput is 0.5. A latency of 4 means that it takes 4 cycles until the result is available. A throughput of 0.5
-means that within one cycle 2 instructions from the intrinsic can be started. In general there are two units on the CPU to
-compute such functions. Because there are 2 units the throughput is 0.5.
-Within one cycle two add intrinsic functions can be started. It takes up to 4 cycles until a result is available.
+Latency and throughput can vary depending on the underlying CPU. For the add intrinsic latency varies from two to four
+and the throughput is 0.5. A latency of four means that it takes four cycles until the result is available. A throughput of 0.5
+means that within one cycle two instructions from the intrinsic can be started. 
+\\
+In general there are two units on the CPU to
+compute such functions. Because there are two units the throughput is 0.5.
+Within one cycle two add intrinsic functions can be started. It takes up to four cycles until a result is available.
 % Software pipelining
@@
 \subsection{How do modern processors realize instruction-level parallelism?}
-Instruction-level parallelism (superscalar) refers to several instruction which are being evaluated simultaneously on
+Instruction-level parallelism (superscalar) refers to several instructions which are being evaluated simultaneously on
 different functional units.
@@
 It can be used for scalar or vector operations.
-The body of the loop is replicated and the logic is adjusted. The loop is unrolled by a factor k, 
-which determines the number of times the body is replicated.
+The body of the loop is replicated and the logic is adjusted accordingly. 
+The loop is unrolled by a factor k, which determines the number of times the body is replicated.
+\\
 Loop unrolling may lead to an incresed demand for registers depending on the factor used. 
-It can increase compile time, the program size because the body is replicated and loop unrolling needs additional space in the instruction cache.
-Loop unrolling is more affective when using vector instructions to exploit instruction-level parallelism.
+It can increase compile time and the program size because the body is replicated and additional space in the instruction cache is needed.
+Loop unrolling is more effective when using vector instructions to exploit instruction-level parallelism.
 
@@
 IPC (instructions per cycle) is an aspect of the processors's performance referring to the average number of instructions
-executed in one clock cylce. It is one indication for a systems performance. 
+executed in one clock cylce. It an indication for the systems performance. 
 A high IPC indicates that the CPU is efficiently used.
@@
 The CPU can access the data directly through the cache.
+\\
 Spatial locality means that elements should be stored efficiently. 
@@
 Everything that is loaded should be used as efficiently as possible.
+\\
 Temporal locality means that programs should be written in such a way that the data is used as often as possible.
@@
 \subsection{What are the differences between data-oriented design and object-oriented design?}
-Data-oriented design (DOD) is a way to design programs that make efficient use of CPU cache. 
+Data-oriented design (DOD) is a way to design programs that make efficient use of the CPU cache. 
 DOD modifies the input to obtain the desired output.
-In object-oriented design everything revolves around the defining, producing and operating on objects.
+In object-oriented design everything revolves around defining, producing and operating on objects.
 Objects interact with each other through functions.
-In data-oriented design functionality and data is seperated. Functions are general purpose,
+In data-oriented design functionality and data is seperated. Functions are general purpose.
 It shifts the perspective onto the data and how it is laid out in memory and read from there.
@@
 \subsection{What are streaming stores?}
-Streaming stores are a way to write data from the CPU directly to random access memory (RAM) and not not through cache.
-Data is not written to cache.
+Streaming stores are a way to write data from the CPU directly to random access memory (RAM) and not through cache.
 It can lead to better performance, if the data is not used.
@@
 L2 is the unified cache for data and instructions. \\
-Outside of the CPU cores there is a L3 cache, which is shared by all cores and represents the interface to main memory.
+Outside of the CPU cores there is an L3 cache, which is shared by all cores and represents the interface to main memory.
 The caches size goes from the smallest L1 to the largest L3 cache.
@@
 Each cache can be thought of as a hash table with for example 64 rows. Each row has a number of buckets,
-where each bucket is a cache line with a size of 64 bytes for Intel CPUs.
+where each bucket is a cache line. The size for a cache line for Intel CPUs is 64 bytes.
 
@@
 	\item g - generates debug information
-	\item fsanitize=address - decets out-of-bounds-access, use-after-free and memory leaks
+	\item fsanitize=address - detects out-of-bounds-access, use-after-free and memory leaks
 	\item fsanitize=undefined - detects undefined behavior at runtime
@@
 \subsection{How could Intel oneAPI help you write better programs?}
-Intel oneAPI is a toolkit containing Intel compilers supporting CPUs,GPUs, FPGAs and other accelerators.
+Intel oneAPI is a toolkit containing Intel compilers supporting CPUs, GPUs, FPGAs and other accelerators.
 It provides a cross-architecture programming model including performance libraries, analyzer and debugger tools as well as 
@@
 It is better to first profile and find bottlenecks that can be optimized.
-The approach shouldn't be to invasive especially in a program written by others.
+The approach shouldn't be too invasive especially in a program written by others.
 It is better to keep things simple.
@@
 % cython: wraparound = False
-% cytho: cdivision = True
+% cython: cdivision = True
 
@@
 \subsection{What are extension types in the context of Python?}
-Extension types in can be used to wrap C / C++ code looking like Python objects.
+Extension types can be used to wrap C / C++ code looking like Python objects.
 They can be used to boost performance. Cython makes it easy to create extension types.
@@
 \subsection{Give a simple description of how to wrap C / C++ code in cython?}
-We have a C / C++ library we want to use in Pythong. 
+We have a C / C++ library we want to use in Python. 
 In our Python code we include the library using \#include "pi.h".
@@
 In SSDs cells are used to store digital bits. 
-Cells are NAND based and can store from one to four bits depending on the type of NAND.
+Cells are NAND-based and can store from one to four bits depending on the type of NAND.
 A group of cells forms a page. The smallest unit to read or write with SSDs is a page.
@@
 NVMe protocols are more efficiently with SATA interface.
-The M.2 form factor exists for PCIe and SATA interfaces.
+The M.2 form factor exists for both, PCIe and SATA interfaces.
 It has multiple key slots, which have to be checked for compatibility.

### 2023-02-19 07_Code/thread_simd/cmake-build-release-gcc/CMakeFiles/thread_simd.dir/compiler_depend.ts

@@ -1,0 +1,2 @@
+# CMAKE generated file: DO NOT EDIT!
+# Timestamp file for compiler generated dependencies management for thread_simd.

### 2023-02-19 08_Code/avx2quickselect/avx2quickselect.cpp

@@ -1,0 +1,55 @@
+#include "avx2sort.h"
+#include <functional>
+#include <random>
+#include <vector>
+
+/*
+ * see https://github.com/viktorleis/perfevent
+ * for installation and usage
+ */
+#include "PerfEvent.hpp"
+
+using namespace std;
+
+/* this code computes performance indicators for std::nth_element and vectorized
+ * quickselect */
+
+int main() {
+  int n = 100000000;
+  int k = n / 2;
+
+  /* create a vector with random integers */
+  vector<int> a(n);
+  auto rand_int = bind(uniform_int_distribution<int>{INT32_MIN, INT32_MAX},
+                       default_random_engine{std::random_device{}()});
+  for (int i = 0; i < n; ++i) {
+    a[i] = rand_int();
+  }
+  auto b = a;
+
+  BenchmarkParameters params;
+  params.setParam("n", to_string(n));
+
+  /* std::nth_element */
+  {
+    params.setParam("_name", " std::nth_element");
+    PerfEventBlock e(n, params, true);
+    std::nth_element(begin(a), begin(a) + k, end(a));
+  }
+
+  /* vectorized quickselect */
+  {
+    {
+      params.setParam("_name", "avx2::quickselect");
+      PerfEventBlock e(n, params, false);
+      avx2::quickselect(b.data(), n, k);
+    }
+  }
+
+  avx2::quicksort(a.data(), k);
+  avx2::quicksort(b.data(), k);
+  avx2::quicksort(a.data() + k + 1, n - k - 1);
+  avx2::quicksort(b.data() + k + 1, n - k - 1);
+  if(a != b || a[k] != b[k])
+    cerr << "failed!";
+}

### 2023-02-19 08_Code/avx2sort/PerfEvent.hpp

@@ -1,0 +1,269 @@
+/*
+
+Copyright (c) 2018 Viktor Leis
+
+Permission is hereby granted, free of charge, to any person obtaining
+a copy of this software and associated documentation files (the
+"Software"), to deal in the Software without restriction, including
+without limitation the rights to use, copy, modify, merge, publish,
+distribute, sublicense, and/or sell copies of the Software, and to
+permit persons to whom the Software is furnished to do so, subject to
+the following conditions:
+
+The above copyright notice and this permission notice shall be
+included in all copies or substantial portions of the Software.
+
+THE SOFTWARE IS PROVIDED "AS IS", WITHOUT WARRANTY OF ANY KIND,
+EXPRESS OR IMPLIED, INCLUDING BUT NOT LIMITED TO THE WARRANTIES OF
+MERCHANTABILITY, FITNESS FOR A PARTICULAR PURPOSE AND
+NONINFRINGEMENT. IN NO EVENT SHALL THE AUTHORS OR COPYRIGHT HOLDERS BE
+LIABLE FOR ANY CLAIM, DAMAGES OR OTHER LIABILITY, WHETHER IN AN ACTION
+OF CONTRACT, TORT OR OTHERWISE, ARISING FROM, OUT OF OR IN CONNECTION
+WITH THE SOFTWARE OR THE USE OR OTHER DEALINGS IN THE SOFTWARE.
+
+ */
+
+#pragma once
+
+#if defined(__linux__)
+
+#include <chrono>
+#include <cstring>
+#include <iomanip>
+#include <iostream>
+#include <map>
+#include <string>
+#include <vector>
+#include <sstream>
+
+#include <asm/unistd.h>
+#include <linux/perf_event.h>
+#include <sys/ioctl.h>
+#include <unistd.h>
+
+struct PerfEvent {
+
+   struct event {
+      struct read_format {
+         uint64_t value;
+         uint64_t time_enabled;
+         uint64_t time_running;
+         uint64_t id;
+      };
+
+      perf_event_attr pe;
+      int fd;
+      read_format prev;
+      read_format data;
+
+      double readCounter() {
+         double multiplexingCorrection = static_cast<double>(data.time_enabled - prev.time_enabled) / (data.time_running - prev.time_running);
+         return (data.value - prev.value) * multiplexingCorrection;
+      }
+   };
+
+   std::vector<event> events;
+   std::vector<std::string> names;
+   std::chrono::time_point<std::chrono::steady_clock> startTime;
[…]

### 2023-02-19 08_Code/avx2sort/avx2sort.cpp

@@ -1,0 +1,69 @@
+#include "avx2sort.h"
+#include <functional>
+#include <random>
+#include <vector>
+
+#if defined(__INTEL_COMPILER)
+#include <ipp.h>
+#endif
+
+/*
+ * see https://github.com/viktorleis/perfevent
+ * for installation and usage
+ */
+#include "PerfEvent.hpp"
+
+using namespace std;
+
+/* this code computes performance indicators for std::sort,
+ * vectorized quicksort and IPP radix sort */
+
+int main() {
+  int n = 100000000;
+
+  /* create a vector with random integers */
+  vector<int> a(n);
+  auto rand_int = bind(uniform_int_distribution<int>{INT32_MIN, INT32_MAX}, default_random_engine{std::random_device{}()});
+  for (int i = 0; i < n; ++i) {
+    a[i] = rand_int();
+  }
+  auto b = a;
+
+  BenchmarkParameters params;
+  params.setParam("n",to_string(n));
+
+  /* std::sort */
+  {
+    params.setParam("_name"," std::sort");
+    PerfEventBlock e(n, params, true);
+    std::sort(begin(b), end(b));
+  }
+
+  /* vectorized quicksort */
+  {
+    auto c = a;
+    {
+      params.setParam("_name","avx2::sort");
+      PerfEventBlock e(n, params, false);
+      avx2::quicksort(c.data(), n);
+    }
+    if(c != b) puts("vectorized quicksort failed!\n");
+  }
+
+#if defined(__INTEL_COMPILER)
+  /* IPP radix sort */
+  {
+    IppSizeL sz;
+    Ipp8u *buffer;
+    ippsSortRadixGetBufferSize_L(n, ipp32s, &sz);
+    buffer = (Ipp8u*) malloc(sz);
+    {
+      params.setParam("_name"," IPP radix");
+      PerfEventBlock e(n, params, false);
+      ippsSortRadixAscend_32s_I_L(a.data(), n, buffer);
+    }
+    free(buffer);
+    if(a != b) puts("vectorized quicksort failed!\n");
+  }
+#endif
+}

### 2023-02-19 08_Code/avx2sort/cmake-build-release-gcc/CMakeFiles/avx2sort.dir/compiler_depend.ts

@@ -1,0 +1,2 @@
+# CMAKE generated file: DO NOT EDIT!
+# Timestamp file for compiler generated dependencies management for avx2sort.

### 2023-02-19 08_Code/fma_shuffle/fma_shuffle.cpp

@@ -1,0 +1,137 @@
+#include "aligned_allocator.h"
+#include <array>
+#include <immintrin.h>
+#include <iostream>
+#include <numeric>
+#include <omp.h>
+#include <vector>
+
+template<class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+using namespace std;
+
+#if defined(_WIN32) // restrict pointers on windows
+#if defined(_MSC_VER) || defined(__ICL)
+#define __restrict__ __restrict
+#endif
+#endif
+
+#define PRAGMA(X) _Pragma(#X)
+#if defined(__INTEL_COMPILER)
+#define __unroll(X) PRAGMA(unroll(X))
+#elif defined(__clang__)
+#define __unroll(X) PRAGMA(clang loop unroll_count(X))
+#elif defined(__GNUC__) || defined(__GNUG__)
+#define __unroll(X) PRAGMA(GCC unroll(X))
+#else
+// do nothing
+#define __unroll(X)
+// define here __unroll macro for your compiler
+#endif
+
+void fma_and_swap(const float *__restrict__ a, const float *__restrict__ b,
+                  const float *__restrict__ c, float *__restrict__ d,
+                  const int n) {
+  for (int i = 0; i < n; i += 2) {
+    d[i + 1] = a[i] * b[i] + c[i]; // fma
+    d[i] = a[i + 1] * b[i + 1] + c[i + 1]; // fma
+  }
+}
+
+
+void fma_and_swap_tuned(const float *__restrict__ a,
+                        const float *__restrict__ b,
+                        const float *__restrict__ c, float *__restrict__ d,
+                        const int n) {
+  // slower with two separate for loops
+  // __unroll is a custom macro for unrolling loops
+  //  __unroll(64) for (int i = 0; i < n; ++i) {
+  //    d[i] = a[i] * b[i] + c[i];
+  //  }
+  //  __unroll(32) for (int i = 0; i < n; i += 2) {
+  //    std::swap(d[i], d[i + 1]);
+  //  }
+
+  //__unroll is a custom macro for unrolling loops
+  __unroll(8) for (int i = 0; i < n; i += 8) {
+#pragma omp simd aligned(a, b, c, d : 64)
+    for (int j = i; j < i + 8; ++j) { 
+      d[j] = a[j] * b[j] + c[j]; // 8 floats (AVX2)
+    }
+#pragma omp simd aligned(d : 64)
[…]

### 2023-02-19 08_Code/transposition8x8/transposition8x8.cpp

@@ -1,0 +1,95 @@
+#include "aligned_allocator.h"
+#include <immintrin.h>
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+template <class T>
+using aligned_vector = std::vector<T, alligned_allocator<T, 64>>;
+
+inline void transpose_8x8_int_naive(int *a) {
+  for (int i = 0; i < 8; ++i) {
+    for (int j = i + 1; j < 8; ++j) {
+      std::swap(a[i * 8 + j], a[j * 8 + i]);
+    }
+  }
+}
+
+// transpose 8 x 8 matrix with intrinsics
+inline void transpose_8x8_int_intrinsics(__m256i *vecs) {
+  __m256 *v = reinterpret_cast<__m256 *>(vecs);
+  __m256 a = _mm256_unpacklo_ps(v[0], v[1]);
+  __m256 b = _mm256_unpackhi_ps(v[0], v[1]);
+  __m256 c = _mm256_unpacklo_ps(v[2], v[3]);
+  __m256 d = _mm256_unpackhi_ps(v[2], v[3]);
+  __m256 e = _mm256_unpacklo_ps(v[4], v[5]);
+  __m256 f = _mm256_unpackhi_ps(v[4], v[5]);
+  __m256 g = _mm256_unpacklo_ps(v[6], v[7]);
+  __m256 h = _mm256_unpackhi_ps(v[6], v[7]);
+  auto tmp = _mm256_shuffle_ps(a, c, 0x4E);
+  a = _mm256_blend_ps(a, tmp, 0xCC);
+  c = _mm256_blend_ps(c, tmp, 0x33);
+  tmp = _mm256_shuffle_ps(b, d, 0x4E);
+  b = _mm256_blend_ps(b, tmp, 0xCC);
+  d = _mm256_blend_ps(d, tmp, 0x33);
+  tmp = _mm256_shuffle_ps(e, g, 0x4E);
+  e = _mm256_blend_ps(e, tmp, 0xCC);
+  g = _mm256_blend_ps(g, tmp, 0x33);
+  tmp = _mm256_shuffle_ps(f, h, 0x4E);
+  f = _mm256_blend_ps(f, tmp, 0xCC);
+  h = _mm256_blend_ps(h, tmp, 0x33);
+  v[0] = _mm256_permute2f128_ps(a, e, 0x20);
+  v[1] = _mm256_permute2f128_ps(c, g, 0x20);
[…]

### 2023-02-19 09_Code/cache_conflicts.cpp

@@ -1,0 +1,32 @@
+#include <iostream>
+#include <omp.h>
+#include <vector>
+
+void dummy(int increment) {
+  std::vector<int> v(100 * increment);
+  // repeatedly updating on 100 different cache lines one value
+  for (int i = 0; i < 10000000; ++i) {
+    for (int j = 0; j < v.size(); j+=increment) {
+      ++v[j];
+    }
+  }
+}
+
+int main() {
+  using namespace std;
+  {
+    auto start = omp_get_wtime();
+    dummy(1020);
+    cout << "increment 1020: " << omp_get_wtime() - start << " seconds" << endl;
+  }
+  {
+    auto start = omp_get_wtime();
+    dummy(1024); // increment leads to L1 cache conflicts
+    cout << "increment 1024: " << omp_get_wtime() - start << " seconds" << endl;
+  }
+  {
+    auto start = omp_get_wtime();
+    dummy(1028);
+    cout << "increment 1028: " << omp_get_wtime() - start << " seconds" << endl;
+  }
+}

### 2023-02-19 10_Code/buggy/buggy.cpp

@@ -1,0 +1,14 @@
+#include <iostream>
+
+int main() {
+  int* dummy = new int[2];
+#pragma omp parallel num_threads(16)
+  {
+    // use Intel Inspector to detect race condition
+    // fsanitize=undefined detects integer overflow
+    dummy[0] += rand();
+  }
+  // fsanitize=address detects out-of-bounds access
+  std::cout << dummy[2];
+  // fsanitize=address detects memory leak
+}

### 2023-02-19 10_Code/buggy/cmake-build-debug-intel/CMakeFiles/buggy.dir/compiler_depend.ts

@@ -1,0 +1,2 @@
+# CMAKE generated file: DO NOT EDIT!
+# Timestamp file for compiler generated dependencies management for buggy.

### 2023-02-19 10_Code/debugger_demo.cpp

@@ -1,0 +1,11 @@
+#include <iostream>
+#include <vector>
+
+int main() {
+  std::vector<int> v(1000, 7);
+  int64_t sum = 0;
+  for (int i = 0; i < v.size(); ++i) {
+    sum += v[i];
+  }
+  std::cout << sum << std::endl;
+}

### 2023-02-19 10_Code/jacobi.cpp

@@ -1,0 +1,211 @@
+//
+// Implementation of the iterative Jacobi method.
+//
+// Given a known, diagonally dominant matrix A and a known vector b, we aim to
+// to find the vector x that satisfies the following equation:
+//
+//     Ax = b
+//
+// We first split the matrix A into the diagonal D and the remainder R:
+//
+//     (D + R)x = b
+//
+// We then rearrange to form an iterative solution:
+//
+//     x' = (b - Rx) / D
+//
+// More information:
+// -> https://en.wikipedia.org/wiki/Jacobi_method
+//
+
+// original source: https://github.com/UoB-HPC/intro-hpc-jacobi/
+
+#include <math.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <sys/time.h>
+
+static int N;
+static int MAX_ITERATIONS;
+static int SEED;
+static double CONVERGENCE_THRESHOLD;
+
+#define SEPARATOR "------------------------------------\n"
+
+// Return the current time in seconds since the Epoch
+double get_timestamp();
+
+// Parse command line arguments to set solver parameters
+void parse_arguments(int argc, char *argv[]);
+
+// Run the Jacobi solver
+// Returns the number of iterations performed
+int run(double *A, double *b, double *x, double *xtmp) {
+  int itr;
+  int row, col;
+  double dot;
+  double diff;
+  double sqdiff;
+  double *ptrtmp;
+
+  // Loop until converged or maximum iterations reached
+  itr = 0;
+  do {
+    // Perfom Jacobi iteration
+    for (row = 0; row < N; row++) {
+      dot = 0.0;
+      for (col = 0; col < N; col++) {
+        if (row != col)
+          dot += A[row + col * N] * x[col];
+      }
+      xtmp[row] = (b[row] - dot) / A[row + row * N];
+    }
+
+    // Swap pointers
+    ptrtmp = x;
+    x = xtmp;
+    xtmp = ptrtmp;
+
+    // Check for convergence
+    sqdiff = 0.0;
+    for (row = 0; row < N; row++) {
+      diff = xtmp[row] - x[row];
+      sqdiff += diff * diff;
+    }
+
+    itr++;
+  } while ((itr < MAX_ITERATIONS) && (sqrt(sqdiff) > CONVERGENCE_THRESHOLD));
+
+  return itr;
+}
+
[…]

### 2023-02-19 10_Code/jacobi_fast.cpp

@@ -1,0 +1,220 @@
+//
+// Implementation of the iterative Jacobi method.
+//
+// Given a known, diagonally dominant matrix A and a known vector b, we aim to
+// to find the vector x that satisfies the following equation:
+//
+//     Ax = b
+//
+// We first split the matrix A into the diagonal D and the remainder R:
+//
+//     (D + R)x = b
+//
+// We then rearrange to form an iterative solution:
+//
+//     x' = (b - Rx) / D
+//
+// More information:
+// -> https://en.wikipedia.org/wiki/Jacobi_method
+//
+
+#include <math.h>
+#include <stdio.h>
+#include <stdlib.h>
+#include <string.h>
+#include <sys/time.h>
+#include <immintrin.h>
+
+static int N;
+static int MAX_ITERATIONS;
+static int SEED;
+static double CONVERGENCE_THRESHOLD;
+
+#define SEPARATOR "------------------------------------\n"
+
+// Return the current time in seconds since the Epoch
+double get_timestamp();
+
+// Parse command line arguments to set solver parameters
+void parse_arguments(int argc, char *argv[]);
+
+// Run the Jacobi solver
+// Returns the number of iterations performed
+int run(double * __restrict__ A, double * __restrict__ b, double * __restrict__ x, double * __restrict__ xtmp) {
+  int itr;
+  int row, col;
+  double dot;
+  double diff;
+  double sqdiff;
+  double *ptrtmp;
+
+  // Loop until converged or maximum iterations reached
+  itr = 0;
+  do {
+    // Perfom Jacobi iteration
+    sqdiff = 0.0;
+// #pragma omp parallel for num_threads(2)
+    for (row = 0; row < N; row++) {
+      dot = 0.0;
+      for (col = 0; col < N; col++) {
+          dot += A[row * N + col] * x[col];
+      }
+      auto new_x = (b[row] - (dot- A[row + row * N] * x[row])) / A[row + row * N];
+      diff = new_x - x[row];
+      sqdiff += diff * diff;
+      x[row] = new_x;
+    }
+
+//    // Swap pointers
+//    ptrtmp = x;
+//    x = xtmp;
+//    xtmp = ptrtmp;
+//
+//    // Check for convergence
+//    sqdiff = 0.0;
+//    for (row = 0; row < N; row++) {
[…]

### 2023-02-19 10_Code/perf_demo/perf_demo.cpp

@@ -1,0 +1,61 @@
+#include <immintrin.h>
+#include <iostream>
+
+/*
+ * You need to install perf first
+ *
+ * then see https://github.com/viktorleis/perfevent
+ * for installation and usage of perfEvent
+ *
+ * You may need to run "sudo sysctl -w kernel.perf_event_paranoid=-1"
+ * and/or add "kernel.perf_event_paranoid = -1" to /etc/sysctl.conf
+ */
+#include "PerfEvent.hpp"
+
+using namespace std;
+
+double sum_matrix_columnwise(const double *mat, int64_t n) {
+  double result = 0;
+  for (int64_t i = 0; i < n; ++i) {
+    for (int64_t j = 0; j < n; ++j) {
+      result += mat[i + j * n];
+    }
+  }
+  volatile double trigger = result;
+  return trigger;
+}
+
+double sum_matrix_rowwise(const double *mat, int64_t n) {
+  double result = 0;
+  for (int64_t i = 0; i < n; ++i) {
+    for (int64_t j = 0; j < n; ++j) {
+      result += mat[i * n + j];
+    }
+  }
+  volatile double trigger = result;
+  return trigger;
+}
+
+int main() {
+  int64_t n = 20000;
+  auto mat = (double *)_mm_malloc(n * n * sizeof(double), 128);
+  for (int64_t i = 0; i < n * n; ++i) {
+    mat[i] = 1.0;
+  }
+
+  BenchmarkParameters params;
+
+  {
+    params.setParam("_name", "columnwise"); // set parameter
+    PerfEventBlock e(n * n, params, true); // start counter
+    sum_matrix_columnwise(mat, n); // computation to benchmark
+  }
+
+  {
+    params.setParam("_name", "   rowwise"); // change parameter
+    PerfEventBlock e(n * n, params, false); // start counter
+    sum_matrix_rowwise(mat, n); // computation to benchmark
+  }
+
+  _mm_free(mat);
+}

### 2023-02-19 11_Code/pi/pure_python/compute_pi_with_python.py

@@ -1,0 +1,22 @@
+import time
+
+
+def compute_pi(num_steps):
+    width = 1.0 / num_steps  # width of a rectangle
+    csum = 0.0  # for summing up all heights of rectangles
+
+    for i in range(num_steps):
+        x = (i + 0.5) * width  # midpoint
+        csum = csum + (1.0 / (1.0 + x * x))
+
+    return csum * 4 * width
+
+
+num_steps = 10000000  # number of rectangles
+
+start_time = time.time()
+pi = compute_pi(num_steps)
+run_time = time.time() - start_time
+
+print("pi with", num_steps, "steps is",
+      pi, "in", "%.6f" % run_time, "seconds\n")

### 2023-02-19 11_Code/pi/with_cython/compute_pi.pyx

@@ -1,0 +1,16 @@
+# cython: language_level = 3
+
+from libc.stdint cimport uint64_t
+
+
+def compute_pi(uint64_t num_steps):
+    cdef double width = 1.0 / num_steps  # width of a rectangle
+    cdef double csum = 0.0  # for summing up all heights of rectangles
+
+    cdef uint64_t i
+    cdef double x
+    for i in range(num_steps):
+        x = (i + 0.5) * width  # midpoint
+        csum = csum + (1.0 / (1.0 + x * x))
+
+    return csum * 4 * width

### 2023-02-19 11_Code/pi/with_cython/compute_pi_with_cython.py

@@ -1,0 +1,12 @@
+import time
+from compute_pi import compute_pi
+
+
+num_steps = 10000000  # number of rectangles
+
+start_time = time.time()
+pi = compute_pi(num_steps) # 30 times faster
+run_time = time.time() - start_time
+
+print("pi with", num_steps, "steps is",
+      pi, "in", "%.6f" % run_time, "seconds\n")

### 2023-02-19 11_Code/scan_enhancer/pure_python/scan_enhancer.py

@@ -1,0 +1,41 @@
+from PIL import Image
+import numpy as np
+from scipy.ndimage import gaussian_filter
+
+
+def rgb2gray(rgb_img):
+    return np.dot(rgb_img, [0.2989, 0.5870, 0.1140])
+
+
+def custom_filter(blurred_img, gray_img):
+    for i in range(gray_img.shape[0]):
+        for j in range(gray_img.shape[1]):
+            d = gray_img[i][j] / (blurred_img[i][j] + 0.00000000001)
+            m = min(255, d * 255 + 0.00000000001)
+            gamma = 0.5 * (m / 255.0) * m * m / 65025
+            m = ((m / 255.0) ** (1 / (gamma + 0.00000000001))) * 255.0
+            gray_img[i][j] = m
+
+
+def enhance_gray_img(gray_img):
+    gray_img_copy = np.copy(gray_img)
+    blurred_img = gaussian_filter(img, sigma=30)
+
+    print(blurred_img.dtype)
+    print(blurred_img.shape)
+    print(gray_img_copy.dtype)
+    print(gray_img_copy.shape)
+
+    custom_filter(blurred_img, gray_img_copy)
+    return gray_img_copy
+
+
+input_img = 'in.jpg'
+output_img = 'out.png'
+
+img = Image.open(input_img)
+img = rgb2gray(np.array(img))
+img = enhance_gray_img(img)
+I8 = img.astype(np.uint8)
+img = Image.fromarray(I8)
+img.save(output_img)

### 2023-02-19 11_Code/scan_enhancer/with_cython/filter.pyx

@@ -1,0 +1,20 @@
+### Cython compiler directives ###
+# cython: language_level = 3
+# cython: boundscheck = False
+# cython: wraparound = False
+# cython: cdivision = True
+##################################
+
+
+# zwei numpy arrays
+def custom_filter(double[:,:] blurred_img, double[:,:] gray_img):
+    cdef:
+        int i, j
+        double d, m, gamma
+    for i in range(gray_img.shape[0]):
+        for j in range(gray_img.shape[1]):
+            d = gray_img[i][j] / (blurred_img[i][j] + 0.00000000001)
+            m = min(255, d * 255 + 0.00000000001)
+            gamma = 0.5 * (m / 255.0) * m * m / 65025
+            m = ((m / 255.0) ** (1 / (gamma + 0.00000000001))) * 255.0
+            gray_img[i][j] = m

### 2023-02-19 11_Code/scan_enhancer/with_cython/scan_enhancer_pyximport.py

@@ -1,0 +1,37 @@
+import pyximport
+
+pyximport.install()
+
+from PIL import Image
+import numpy as np
+from scipy.ndimage import gaussian_filter
+#pyx importieren
+from filter import custom_filter
+import time
+
+
+def rgb2gray(rgb_img):
+    # return np.asarray(rgb_img, dtype=np.double)
+    return np.dot(rgb_img, [0.2989, 0.5870, 0.1140])
+
+
+def enhance_gray_img(gray_img):
+    gray_img_copy = np.copy(gray_img)
+    blurred_img = gaussian_filter(img, sigma=30)
+    start_time = time.time()
+
+    custom_filter(blurred_img, gray_img_copy)
+    run_time = time.time() - start_time
+    print("custom_filter", "in", "%.6f" % run_time, "seconds\n")
+    return gray_img_copy
+
+
+input_img = 'in.jpg'
+output_img = 'out.png'
+
+img = Image.open(input_img)
+img = rgb2gray(np.array(img))
+img = enhance_gray_img(img)
+I8 = img.astype(np.uint8)
+img = Image.fromarray(I8)
+img.save(output_img)

### 2023-02-19 12_Code/book_example/book.pyx

@@ -1,0 +1,9 @@
+# cython: language_level = 3
+
+# cdef class statement tells cython to make an extension type
+cdef class Cy_Book: # Cy_Book is an extension type
+    # either "readonly" or "public"
+    cdef readonly str title # data field
+
+    def __init__(self, title):
+        self.title = title

### 2023-02-19 12_Code/book_example/book_demo.py

@@ -1,0 +1,10 @@
+# pyximport automatically recompiles and reloads your .pyx modules
+import pyximport
+pyximport.install()
+
+# import your extension type
+from book import Cy_Book
+
+# create an object of extension type Cy_Book
+some_book = Cy_Book('Cython: A Guide for Python Programmers')
+print(some_book.title)

### 2023-02-19 12_Code/sampler/DSampling.pyx

@@ -1,0 +1,111 @@
+# cython: language_level = 3
+# cython: boundscheck = False
+# cython: wraparound = False
+# cython: embedsignature=True
+
+from libc.stdint cimport uint64_t, uint32_t
+from libc.stdlib cimport malloc, free
+from libc.stdlib cimport qsort
+
+cdef inline uint64_t rotl(uint64_t x, int k):
+    return (x << k) | (x >> (64 - k))
+
+cdef inline uint64_t next_rand(uint64_t* s0, uint64_t* s1):
+    s1[0] = s1[0] ^ s0[0]
+    s0[0] = rotl(s0[0], 24) ^ s1[0] ^ (s1[0] << 16)
+    s1[0] = rotl(s1[0], 37)
+    return s0[0] + s1[0]
+
+cdef inline double next_double_rand(uint64_t* s0, uint64_t* s1):
+    cdef uint64_t n = (next_rand(s0, s1) >> 12) | 0x3FF0000000000000
+    cdef double* num = <double*>&n
+    return num[0] - 1.0
+    
+cdef class Rand:
+    cdef:
+        uint64_t s1 
+        uint64_t s2
+        
+    def __init__(self, uint32_t seed = 0):
+        self.s1 = 1314472907419283471ull + <uint64_t>seed # seed 1
+        self.s2 = 7870872464127966567ull + <uint64_t>seed # seed 2
+        cdef int i
+        for i in range(10): # warm up
+            self.next_double()
+        
+    cpdef double next_double(self):
+        return next_double_rand(&self.s1, &self.s2)
+     
+cdef struct Item:
+    double key
+    int index
+    
+cdef int comparator(const void* pa, const void* pb) nogil:
+    if (<Item*>pa)[0].key < (<Item*>pb)[0].key:
+        return 1
+    elif (<Item*>pa)[0].key > (<Item*>pb)[0].key:
+        return -1
+    else:
+        return 0
+
+cdef class Sampler:
+    """
+    Sampler for discrete distributions.
+    """
+    cdef:
+        Rand generator
+        double *cprobs
+        int* categories
+        int length
+    
+    def __cinit__(self, double[:] cprobs, uint32_t seed = 0):
+        self.length = cprobs.shape[0]
+        if(self.length < 2):
+            raise Exception("Less than two categories given.")
+        cdef int i
+        cdef csum = 0.0
+        cdef bint less_zero = False
+        for i in range(self.length):
+            csum = csum + cprobs[i]
[…]

### 2023-02-19 12_Code/sampler/how_to_use.py

@@ -1,0 +1,32 @@
+import pyximport
+pyximport.install()
+# you can also compile the module with cythonize -i DSampling.pyx
+# in that case you don't need pyximport
+
+from DSampling import Sampler
+import numpy as np
+
+# given the following probabilities, we sample the corresponding indices
+probabilities = np.array([0.2, 0.1, 0.3, 0.4])
+# normalize --> sum of all probabilities = 1.0, sampler throws error when that's not the case
+probabilities = probabilities / sum(probabilities) 
+
+# provide seed 0 -- 2^32-1, default seed is 0
+# here, seed is random
+seed = np.random.randint(2**32 - 1, dtype=np.uintc)
+
+# construct the Sampler
+sampler = Sampler(probabilities, seed)
+
+# single sample
+print("single sample:", sampler.sample())
+# you sample the index, for example if sample is 2
+# then the thing with 0.3 probability was sampled
+
+# a faster way is to fill an integer array with samples
+arr = np.zeros(10, dtype=np.intc)
+# fill the array with samples
+sampler.fill_with_samples(arr)
+print("samples in array:", arr)
+# you can also fill just a part of an array with samples through slicing
+# for example: sampler.fill_with_samples(arr[:2])

### 2023-02-19 12_Code/sampler/performance_comparison.py

@@ -1,0 +1,55 @@
+import pyximport
+pyximport.install()
+
+import time
+from DSampling import Sampler
+import numpy as np
+
+# distribution of letters in alphabet, https://en.wikipedia.org/wiki/Letter_frequency
+probabilities = np.array([8.167, 1.492, 2.202, 4.253, 12.702, 2.228, 2.015, 6.094, 6.966, 0.153, 1.292, 4.025, 2.406, 6.749, 7.507, 1.929, 0.095, 5.987, 6.327, 9.356, 2.758, 0.978, 2.560, 0.150, 1.994, 0.077])
+# normalize --> sum of all probabilities = 1.0, sampler throws error when that's not the case
+probabilities = probabilities / sum(probabilities) 
+
+# provide seed 0 -- 2^32-1, default seed is 0
+# here, seed is random
+seed = np.random.randint(2**32 - 1, dtype=np.uintc)
+
+# construct the Sampler
+sampler = Sampler(probabilities, seed)
+
+#### performance measurement ####
+n = 1000000  # amount of samples
+arr = np.zeros(n, dtype=np.intc)
+start_time = time.time()
+sampler.fill_with_samples(arr)
+run_time = time.time() - start_time
+print("n =", n)
+print("fill array with samples", "in", "%.4f" % run_time, "seconds")
+
+start_time = time.time()
+for _ in range(n):
+    sampler.sample()
+run_time = time.time() - start_time
+print("sampling by one cython", "in", "%.4f" % run_time, "seconds")
+
+# sampling with python only
+class Py_Sampler:
+    def __init__(self, probs):
+        self.cprobs = np.copy(probs)
+        for i in range(1, probs.shape[0]):
+            self.cprobs[i] = self.cprobs[i - 1] + self.cprobs[i]
+
+    def sample(self):
+        random_double = np.random.random()
+        for i in range(self.cprobs.shape[0] - 1):
+            if(random_double < self.cprobs[i]):
+                return i
+        return self.cprobs.shape[0] - 1
+
+# python sampler is significantly slower then the cython sampler
+py_sampler = Py_Sampler(probabilities)
+start_time = time.time()
+for _ in range(n):
+    py_sampler.sample()
+run_time = time.time() - start_time
+print("sampling by one python", "in", "%.4f" % run_time, "seconds")

### 2023-02-19 12_Code/soft_shrink_grp/example_usage.py

@@ -1,0 +1,13 @@
+from soft.shrink import grp
+import numpy as np
+
+rows = 4000
+columns = 8000
+
+z = np.random.rand(rows, columns)
+tau = 0.5
+Lcum = np.array(list(range(0, rows, 8)))
+
+zr, shrinkednorm = grp(z, tau, Lcum)
+
+print(shrinkednorm)

### 2023-02-19 12_Code/soft_shrink_grp/setup.py

@@ -1,0 +1,25 @@
+from setuptools import setup
+from Cython.Build import cythonize
+import sys
+
+# compile and link flags
+if sys.platform.startswith('win'):
+    # to compile this package on windows you need MSVC
+    compile_args = ['/openmp', '/fp:fast']
+    link_args = ['/openmp']
+elif sys.platform.startswith('darwin'):
+    # use real gcc (like gcc-9) and not clang on mac
+    compile_args = ['-fopenmp', '-ffast-math']
+    link_args = ['-lomp']
+else:
+    # linux
+    compile_args = ['-fopenmp', '-ffast-math']
+    link_args = ['-fopenmp']
+
+setup(
+    name='shrink',
+    version='0.1',
+    ext_modules=cythonize(
+        ['soft/shrink.pyx'],
+        aliases={'XCOMPARGS': compile_args, 'XLINKARGS': link_args}
+    ))

### 2023-02-19 12_Code/soft_shrink_grp/soft/shrink.pyx

@@ -1,0 +1,30 @@
+# distutils: extra_compile_args = XCOMPARGS
+# distutils: extra_link_args = XLINKARGS
+# cython: language_level = 3
+# cython: boundscheck = False
+# cython: wraparound = False
+# cython: cdivision = True
+# cython: embedsignature = True
+
+import numpy as np
+
+from cython.parallel cimport prange
+
+def grp(double[:,::1] z, double tau, long[:] Lcum, long n_threads = 4):
+    """ This function computes ... <give here a description> """
+    cdef double shrinkednorm = 0.0, gnorm, num, fac
+    cdef int i, j, k;
+
+    for i in prange(Lcum.shape[0] - 1, nogil=True, schedule='static', num_threads=n_threads):
+        for j in range(z.shape[1]):
+            gnorm = 0.0
+            for k in range(Lcum[i], Lcum[i + 1]):
+                gnorm += z[k, j] * z[k, j]
+            if gnorm > 0.0:
+                gnorm = gnorm ** 0.5
+                fac = max(0.0, (1.0 - tau / gnorm))
+                for k in range(Lcum[i], Lcum[i + 1]):
+                    z[k, j] = z[k, j] *  fac
+                shrinkednorm += fac * gnorm
+
+    return np.asarray(z), shrinkednorm

### 2023-02-19 13_Code/memmap.py

@@ -1,0 +1,17 @@
+import numpy as np
+
+# Attention, here 16*10^9 bytes are written and then read! Can take a long time!
+# create a float64 200000 * 10000 matrix (16*10^9 bytes) in working directory
+A = np.memmap('A.memmap', dtype='float64', mode='w+', shape=(200000, 10000))
+
+# fill matrix with ones
+A.fill(1)
+
+# print sum of all entries in A (--> 2000000000.0)
+print(np.sum(A))
+
+# flush all changes in the array (guarantees that the data is actually on disc)
+A.flush()
+
+# see here for more information about memory-mapped files in NumPy
+# https://numpy.org/doc/stable/reference/generated/numpy.memmap.html#numpy.memmap

### 2023-02-19 13_Code/small_IO/small_IO.cpp

@@ -1,0 +1,71 @@
+#include <omp.h>
+
+#include <filesystem>
+#include <fstream>
+#include <iostream>
+#include <mutex>
+#include <vector>
+
+using namespace std;
+namespace fs = filesystem;
+
+vector<string> ls_files_recursive(const string& path) {
+  vector<string> files;
+  files.reserve(30000);
+  for (auto& p : fs::recursive_directory_iterator(path)) {
+    if (p.is_regular_file()) {
+      const auto filename = p.path().string();
+      files.push_back(filename);
+    }
+  }
+  return files;
+}
+
+string get_title(const string& file) {
+  const static string tag = "<page title>\": ";
+  ifstream fin(file);
+  string line;
+  while (getline(fin, line)) {
+    auto position = line.find(tag);
+    if (position != string::npos) {
+      position = position + tag.size();
+      // extract string between ': "' and '",' quotes are included
+      line = line.substr(position, line.size() - position - 1);
+      return line;
+    }
+  }
+  return "NO_TITLE";
+}
+
+// extract titles from different files
+// "vector<string> &files" contains the paths to the files
+vector<string> get_titles(const vector<string>& files, int threads) {
+  vector<string> titles(files.size());
+#pragma omp parallel for schedule(dynamic, 8) num_threads(threads)
+  for (size_t i = 0; i < files.size(); ++i) {
+    // function get_title opens file and extracts the title
+    titles[i] = get_title(files[i]);
+  }
+  return titles;  // return a vector of titles
+}
+
+// before running on linux "sync; echo 3 > /proc/sys/vm/drop_caches"
+// --> clears the page cache, dentries, and inodes
+// on OSX "sync && sudo purge"
+int main() {
+  const string camera_specs_dir = "2013_camera_specs";
+  int threads = 16;
+
+  double tic = omp_get_wtime();
+  vector<string> files = ls_files_recursive(camera_specs_dir);
+  double time_in_s_ls = omp_get_wtime() - tic;
+
+  tic = omp_get_wtime();
+  auto titles = get_titles(files, threads);
+  double time_in_s_get_titles = omp_get_wtime() - tic;
+
+  cout << "ls_files_recursive: " << time_in_s_ls << " s" << endl;
+  cout << "get_titles: " << time_in_s_get_titles << " s" << endl;
+  cout << "total: " << time_in_s_ls + time_in_s_get_titles << " s" << endl;
+  cout << "titles count: " << titles.size() << endl;
+}

### 2023-02-19 14_Code/NUMA_data_placement.cpp

@@ -1,0 +1,36 @@
+#include <fstream>
+#include <iostream>
+
+using namespace std;
+
+// create random data (on linux)
+// head -c 1GB /dev/urandom > 1GB_random_data
+
+int main() {
+  ifstream ifs("1GB_random_data", ios::binary | ios::ate);
+  size_t size = ifs.tellg();
+  ifs.seekg(0);
+
+  auto* a = (int*)malloc(size);   // No calloc! No vector!
+  auto* b = (int*)malloc(size);   // No calloc! No vector!
+  size_t n = size / sizeof(int);  // number of integers
+
+  // redundant parallel initialization to place the data upfront
+  // data is placed in the memories closest to the threads executing the loop
+#pragma omp parallel for schedule(static)
+  for (size_t i = 0; i < n; i++) {
+    a[i] = 0;
+  }
+
+  // at this point, the placement has been decided upon already
+  ifs.read((char*)a, size);  // read data to array a
+
+  // each thread processes data from the memory closest to it
+#pragma omp parallel for schedule(static)
+  for (int i = 0; i < n; i++) {
+    b[i] = 2 * a[i];
+  }
+
+  free(a);
+  free(b);
+}

### 2023-02-19 14_Code/mpi_hello.cpp

@@ -1,0 +1,15 @@
+#include <mpi.h>
+#include <iostream>
+
+int main(int argc, char *argv[]) {
+    // using MPI with C++ bindings
+    MPI::Init(argc, argv); // processes are independent until MPI initialization
+    // now processes can collaborate, send/receive messages or synchronize
+    // until reaching MPI::Finalize().
+    int numP = MPI::COMM_WORLD.Get_size(); // get the number of processes
+    int myId = MPI::COMM_WORLD.Get_rank(); // get the ID of the process
+
+    // every process prints Hello
+    std::cout << "Process " << myId << " of " << numP << ": Hello, world!\n";
+    MPI::Finalize(); // terminate MPI
+}

### 2023-02-19 project/main.cpp

@@ -1,0 +1,164 @@
+#include <iostream>
+#include <cstdio>
+#include <algorithm>
+#include <string>
+#include <vector>
+
+using namespace std;
+
+struct Pixel {
+    unsigned char r, g, b;
+};
+
+Pixel median(vector<Pixel> v) {
+    sort(v.begin(), v.end(), [](Pixel a, Pixel b) {
+        return a.r < b.r;
+    });
+    Pixel p = v[v.size() / 2];
+    sort(v.begin(), v.end(), [](Pixel a, Pixel b) {
+        return a.g < b.g;
+    });
+    p.g = v[v.size() / 2].g;
+    sort(v.begin(), v.end(), [](Pixel a, Pixel b) {
+        return a.b < b.b;
+    });
+    p.b = v[v.size() / 2].b;
+    return p;
+}
+
+void medianFilter(string inputFile, string outputFile, int width, int height, int filterSize) {
+    FILE* inFile = fopen(inputFile.c_str(), "rb");
+    FILE* outFile = fopen(outputFile.c_str(), "wb");
+    if (inFile == NULL || outFile == NULL) {
+        printf("Error: Cannot open file\n");
+        return;
+    }
+
+    char magicNumber[3];
+    fscanf(inFile, "%s", magicNumber);
+    int maxVal;
+    fscanf(inFile, "%d", &width);
+    fscanf(inFile, "%d", &height);
+    fscanf(inFile, "%d", &maxVal);
+    fprintf(outFile, "P6\n%d %d\n%d\n", width, height, maxVal);
+
+    int halfSize = filterSize / 2;
+    vector<Pixel> window(filterSize * filterSize);
+
+    for (int i = 0; i < height; ++i) {
+        for (int j = 0; j < width; ++j) {
+            for (int y = -halfSize; y <= halfSize; ++y) {
+                for (int x = -halfSize; x <= halfSize; ++x) {
+                    int ii = i + y;
+                    int jj = j + x;
+                    if (ii < 0) ii = 0;
+                    if (ii >= height) ii = height - 1;
+                    if (jj < 0) jj = 0;
+                    if (jj >= width) jj = width - 1;
+                    fseek(inFile, sizeof(char) * (ii * width + jj) * 3, SEEK_SET);
+                    fread(&window[(y + halfSize) * filterSize + x + halfSize], sizeof(char), 3, inFile);
+                }
+            }
+            Pixel p = median(window);
[…]

### 2023-02-23 Assignments/paper.tex
update

@@
 \subsection{What is Cython?}
-Cython is a superset of Python. 
-It is used for computations that could slow down performance.
-Python code tends to be slow. 
-Cython is used for critical computations to speed up perfromance.
-It is an interface with C and C++ libraries.
+Cython is a programming language. 
+It is a superset of Python used to write C / C++ extensions for Python.
+Cython helps improve performance of Python programs.
 
 \subsection{Describe an approach how Python programs can be accelerated with the help of Cython.}
+Cython can be used in sections of Python programs that are performance critical.
+Python is weak in performance and Cython can be used to improve speed. 
+This can be done by identifying sections in Python code that slow down performance.
+These so called critical sections are outsourced to Cython. 
+In Cython the code sections are written in C / C++ or Cython.
+As Cython is a superset of Python valid syntax in Python is valid in Cython aswell.
+The Cython modules can be imported in a Python program and therefore improve performance.
 
@@
 \subsection{Name and describe two compiler directives in Cython.}
+Compiler directives are set through a special header comment.
+This comment is written before any code.
+Additional flags set in the command line overwrite the ones set as header comments.
+\\
+boundscheck: This directive can be set to true or false and is set to true by default.
+If set to false, Cython assumes that indexing operations won't cause any index errors.
+Conditions that would cause index errors may instead cause segfaults or data corruption.
+\\
+wraparound: This directive is set to true by default. 
+In Python arrays and sequences can be indexed relative to the end.
+Negative indexing is allowed in Python, but not supported in C.
+If the directive wraparound is set to false, Cython can't check or correctly handle negative indices.
+If set to true, negative indexing usually raises an index error.
+For safe usage it is recommended to only use this for code that doesn't process negative indices.
+
 % cython: language_level = 3
@@
 \begin{itemize}
-	\item def - normal python function
+	\item def - is a normal python function
 	\item cdef - can be called only in cython
@@
 \subsection{What are typed memoryviews especially useful for in Cython?}
-Memoryviews are useful for passing Python of Numpy arrays to Cython.
+Typed memoryviews are used to pass Python or Numpy arrays to Cython.
+It provides a view on the underlying memory.
+Typed memoryviews provide access to memory buffers. 
+They are especially useful for those underlying Numpy arrays.
+Memoryviews provide inside on the underlying memory layout, which can be used for code optimization.
 
@@
 \subsection{What is the purpose of garbage collection in SSDs?}
-A block contains pages. The number of invalid pages is compared to a threshold. 
-If it exceeds the threshold, the block is claimed by garbage collection.
-Furthermore, garbage collection moves related data segments next to each other.
+Garbage collection is used to periodically optimize an SSD and maintain performance.
+Garbage collection claims blocks that have more invalid pages than a given threshold.
+The blocks can later be used for write operations.
+It also moves realted data segments next to each other.
 
@@
 War leveling is used to prolong the life of an SSD.
+It should prevent an unbalanced usage of pages, meaning that certain pages are overused and others not at all.
 
@@
 \subsection{Discuss three different recommendations for writing code for SSDs.}
-When writing code for SSDs it is important to avoid scattered updates and use compact data structures.
-Additionally, for large IO, only a few threads suffice to achieve maximum performance.
+\textbf{Compact data structures} should be used when writing code for SSDs. 
+The smallest unit to update in SSDs is a page. 
+Even if only a single bit is updated, it will always result in an SSD write of at least 4 KB.
+The same goes for reading.
+Therefore, small updates should be avoided and compact data structures should be used.
+\\
+Code for SSDs should \textbf{avoid full SSD usage}. 
+This influences the writing performance and write amplification factor.
+A full SSD causes more blocks to be moved around to free a block, which affects performance.
+\\
+An SSD has multiple levels of internal parallelism. 
+For /\textbf{small IO multiple threads} should be used. 
+When reading small pages, multiple threads make efficient use of the internal SSD parallelism.
+For big IO only a few threads suffice, as we can make use of an SSD's internal parallelism.
 
@@
 \subsection{How could you solve problems that do not fit in DRAM without major code adjustments?}
-To solve such problems one can use memory mapped files on flash.
+To solve problems that do not fit in RAM one can use memory mapped files on flash. 
+This can also be done using Numpy for computations that don't fit into memory.
 
@@
 %
-\begin{thebibliography}{8}
-
-\bibitem{ref_url1}
-Armv8-A Reference Manual pdf, pages 30-70, pages 1708-1808 \\ \url{https://developer.arm.com/documentation/ddi0487/ga}.
-
-\end{thebibliography}
+% \begin{thebibliography}{8}
+% 
+% \bibitem{ref_url1}
+% Armv8-A Reference Manual pdf, pages 30-70, pages 1708-1808 \\ \url{https://developer.arm.com/documentation/ddi0487/ga}.
+%
+% \end{thebibliography}
 \end{document}

### 2023-02-23 project/main.cpp

@@ -68,26 +68,79 @@ void medianFilter(string inputFile, string outputFile, int width, int height, in
     fclose(outFile);
 }
 
-void readPPMHeader(FILE *file, unsigned char *header) {
-    fread(header, 15, 1, file);
-}
 
-void readPPMImage(FILE *file, unsigned char *image, int size) {
-    fread(image, size, 1, file);
-}
 
-void writePPM(FILE *file, unsigned char *header, unsigned char *image, int size) {
-    fwrite(header, 15, 1, file);
-    fwrite(image, size, 1, file);
-}
+// Convert PPM image to PGM greyscale image
+int convert_ppm_to_pgm(const char* infile_name, const char* outfile_name) {
 
-void removeRed(unsigned char *image, unsigned char *withoutredimage, int size) {
-    int i;
-    for(i = 0; i < size; i+=3) {
-        withoutredimage[i] = 0; // set red component to 0
-        withoutredimage[i+1] = image[i+1]; // set blue to image
-        withoutredimage[i+2] = image[i+1]; // set green to image
+    // open the input file
+    FILE* infile = fopen(infile_name, "rb");
+    if (!infile) {
+        fprintf(stderr, "Error: could not open input file %s\n", infile_name);
+        return 1;
     }
+
+    // read the magic number
+    char magic_number[3];
+    if (fscanf(infile, "%2s", magic_number) != 1 || magic_number[0] != 'P' || magic_number[1] != '6') {
+        fprintf(stderr, "Error: input file is not in PPM format\n");
+        fclose(infile);
+        return 1;
+    }
+
+    // read the image width, height, and max color value
+    int width, height, maxval;
+    if (fscanf(infile, "%d %d %d\n", &width, &height, &maxval) != 3 || maxval > 255) {
+        fprintf(stderr, "Error: invalid PPM format\n");
+        fclose(infile);
+        return 1;
+    }
+
+    // allocate memory for the RGB pixel data
+    vector<unsigned char> data(width * height * 3);
+
+    // read the pixel data
+    if (fread(data.data(), 1, data.size(), infile) != data.size()) {
+        fprintf(stderr, "Error: could not read pixel data\n");
+        fclose(infile);
+        return 1;
+    }
+
+    // close the input file
+    fclose(infile);
+
+    // open the output file
+    FILE* outfile = fopen(outfile_name, "wb");
+    if (!outfile) {
+        fprintf(stderr, "Error: could not open output file\n");
+        return 1;
[…]

### 2023-02-23 project/otsu.cpp

@@ -1,0 +1,104 @@
+//
+// Created by Sheela on 20/02/23.
+//
+
+#include "otsu.h"
+
+#include <iostream>
+#include <vector>
+#include <cmath>
+#include <cstdio>
+
+using namespace std;
+
+vector<vector<int>> read_pgm(const char *filename) {
+    FILE *infile = fopen(filename, "rb");
+    if(!infile) {
+        fprintf(stderr, "No input file");
+        exit(1);
+    }
+
+    char magic[3], comment[100];
+    int width, height, maxval;
+    fscanf(infile, "%2s\n%100[^\n]\n%d %d\n%d\n", magic, comment, &width, &height, &maxval);
+    vector<vector<int>> image(height, vector<int>(width));
+    fread(image[0].data(), 1, height*width, infile);
+    fclose(infile);
+    return image;
+}
+
+void write_pgm(const vector<vector<int>>& image, const char* filename) {
+    FILE* outfile = fopen(filename, "wb");
+    if (!outfile) {
+        cerr << "Error opening file: " << filename << endl;
+        exit(1);
+    }
+    const int height = image.size();
+    const int width = image[0].size();
+    fprintf(outfile, "P5\n%d %d\n255\n", width, height);
+    fwrite(image[0].data(), 1, height*width, outfile);
+    fclose(outfile);
+}
+
+// Function to compute the Otsu threshold
+int otsu_threshold(const vector<vector<int>>& image) {
+    const int height = image.size();
+    const int width = image[0].size();
+    vector<int> hist(256, 0);
+    for (int i = 0; i < height; i++) {
+        for (int j = 0; j < width; j++) {
+            hist[image[i][j]]++;
+        }
+    }
+    const int total = height * width;
+    double sum = 0.0;
+    for (int i = 0; i < 256; i++) {
+        sum += i * hist[i];
+    }
+    double sumB = 0.0;
+    int wB = 0;
+    int wF = 0;
+    double varMax = 0.0;
+    int threshold = 0;
+    for (int i = 0; i < 256; i++) {
+        wB += hist[i];
+        if (wB == 0) {
+            continue;
+        }
+        wF = total - wB;
+        if (wF == 0) {
+            break;
+        }
+        sumB += i * hist[i];
+        const double mB = sumB / wB;
+        const double mF = (sum - sumB) / wF;
+        const double varBetween = wB * wF * pow(mB - mF, 2);
+        if (varBetween > varMax) {
+            varMax = varBetween;
+            threshold = i;
[…]

### 2023-02-23 project/otsu.h

@@ -1,0 +1,13 @@
+//
+// Created by Sheela on 20/02/23.
+//
+
+#ifndef PROJECT_OTSU_H
+#define PROJECT_OTSU_H
+
+
+class otsu {
+};
+
+
+#endif//PROJECT_OTSU_H
